---
title: "FoodAccessPREP"
author: "Oliver Temel, Mairangi Campbell and Jiayi Lu"
IDs: " , , 300633898" 
date: "2026-09-03"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r 2015 load, echo = FALSE, include = FALSE}
library(readxl)
DATA2015 <- read_excel("FoodAccessResearchAtlasData2015.xlsx", 
                       sheet = 3, na = c("", "NULL"))
```

```{r 2019 load, echo = FALSE, include = FALSE}
DATA2019 <- read_excel("FoodAccessResearchAtlasData2019.xlsx", 
                       sheet = 3, na = c("", "NULL"))
```

```{r, echo = FALSE, include=FALSE}
dim(DATA2015)
dim(DATA2019)
names(DATA2015)
names(DATA2019)
str(DATA2015)
str(DATA2019)
```

```{r, echo = FALSE, include=FALSE}
# To Unify variable names
names(DATA2015) <- tolower(names(DATA2015))
names(DATA2019) <- tolower(names(DATA2019))
```

```{r, include=FALSE}
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
```

```{r, echo = FALSE, include = FALSE}
# Select variables we need
data2015_clean <- DATA2015 %>%
  select(
    censustract, state, county, urban, pop2010, ohu2010, laseniors1, laseniors10, lasnap1, laseniors1share, laseniors10share, lasnap1share, lasnap10share, lasnap10, laseniors20, laseniorshalf, laseniorshalfshare, tractseniors, tractsnap
  )

data2019_clean <- DATA2019 %>%
  select(
    censustract, state, county, urban, pop2010, ohu2010, laseniors1, laseniors10, laseniors1share, laseniors10share, lasnap1share, lasnap10share, lasnap1, lasnap10, laseniors20, laseniorshalf, laseniorshalfshare, tractseniors, tractsnap, 
)
```

```{r, echo = FALSE, include=FALSE}
numeric_vars <- c(
  "urban", "pop2010", "ohu2010", "laseniors1", "laseniors10", "laseniors1share", "laseniors10share", "lasnap1", "lasnap10", "lasnap1share", "lasnap10share", "tractseniors", "tractsnap"
)

data2015_clean <- data2015_clean %>%
  mutate(
    across(
      any_of(numeric_vars),
      as.numeric
    )
  )

data2019_clean <- data2019_clean %>%
  mutate(
    across(
      any_of(numeric_vars),
      as.numeric
    )
  )

combined_data <- bind_rows(
  data2015_clean,
  data2019_clean
)

combined_data
```

```{r, echo = FALSE, include=FALSE}
sapply(
  data2015_clean[numeric_vars],
  class
)

sapply(
  data2019_clean[numeric_vars],
  class
)
```

```{r, echo = FALSE, include = FALSE}
data2015_clean <- data2015_clean %>%
  mutate(year = 2015)

data2019_clean <- data2019_clean %>%
  mutate(year = 2019)
```

```{r Urban / rural Classifier, echo = FALSE}
#Create Urban/Rural
data2015_clean <- data2015_clean %>%
  mutate(
    year = 2015,
    area = if_else(
      urban == 1,
      "Urban",
      "Rural",
      missing = NA_character_
    )
  )

data2019_clean <- data2019_clean %>%
  mutate(
    year = 2019,
    area = if_else(
      urban == 1,
      "Urban",
      "Rural",
      missing = NA_character_
    )
  )
```

```{r, echo = FALSE, include=FALSE}
# 2015 of senior and snap
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_n = case_when(
      urban == 1 ~ laseniors1,
      urban == 0 ~ laseniors10,
      TRUE ~ NA_real_
    ),

    snap_la_n = case_when(
      urban == 1 ~ lasnap1,
      urban == 0 ~ lasnap10,
      TRUE ~ NA_real_
    )
  )

# 2019 of senior and snap
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_n = case_when(
      urban == 1 ~ laseniors1,
      urban == 0 ~ laseniors10,
      TRUE ~ NA_real_
    ),

    snap_la_n = case_when(
      urban == 1 ~ lasnap1,
      urban == 0 ~ lasnap10,
      TRUE ~ NA_real_
    )
  )
```

```{r, echo = FALSE, include=FALSE}
# Show precentage
# 2015
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_pct = case_when(
      urban == 1 ~ laseniors1share * 100,
      urban == 0 ~ laseniors10share * 100,
      TRUE ~ NA_real_
    ),

    snap_la_pct = case_when(
      urban == 1 ~ lasnap1share * 100,
      urban == 0 ~ lasnap10share * 100,
      TRUE ~ NA_real_
    )
  )

# 2019
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_pct = case_when(
      urban == 1 ~ laseniors1share,
      urban == 0 ~ laseniors10share,
      TRUE ~ NA_real_
    ),

    snap_la_pct = case_when(
      urban == 1 ~ lasnap1share,
      urban == 0 ~ lasnap10share,
      TRUE ~ NA_real_
    )
  )
```

```{r, echo = FALSE, include=FALSE}
# Seniors
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_n = if_else(
      urban == 1,
      laseniors1,
      laseniors10
    )
  )

# SNAP
data2015_clean <- data2015_clean %>%
  mutate(
    snap_la_n = if_else(
      urban == 1,
      lasnap1,
      lasnap10
    )
  )

# Seniors
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_n = if_else(
      urban == 1,
      laseniors1,
      laseniors10
    )
  )

# SNAP
data2019_clean <- data2019_clean %>%
  mutate(
    snap_la_n = if_else(
      urban == 1,
      lasnap1,
      lasnap10
    )
  )
```

```{r, echo = FALSE, include=FALSE}
summary(data2015_clean)
summary(data2019_clean)
```

```{r, echo = FALSE, include=FALSE}
#data2019_clean <- data2019_clean %>%
#  mutate(
#    across(
#      c(laseniors1, laseniors10, lasnap1, lasnap10,
#        laseniors20,laseniorshalf,laseniorshalfshare,senior_la_n, snap_la_n),
#      ~ as.numeric(ifelse(.x == "NULL", "0", .x))
#    )
#  )
```

```{r, echo = FALSE, include=FALSE}
summary(data2019_clean)
```

```{r, echo = FALSE, include=FALSE}
library(knitr)

dataset_table <- data.frame(
  Dataset = c("2015 Food Access Research Atlas",
              "2019 Food Access Research Atlas"),
  Observations = c(72864, 72531),
  Variables = c(147, 147),
  Unit_of_Analysis = c("Census tract", "Census tract"),
  Main_Data_Types = c(
    "Geographic identifiers, urban/rural indicator, population counts, food-access measures",
    "Geographic identifiers, urban/rural indicator, population counts, food-access measures"
  )
)

kable(
  dataset_table,
  caption = "Structure of the 2015 and 2019 Food Access Research Atlas datasets",
  col.names = c(
    "Dataset",
    "Observations",
    "Variables",
    "Unit of Analysis",
    "Main Data Types"
  ),
  align = c("l", "r", "r", "l", "l")
)
```

```{r, echo = FALSE, include=FALSE}
summary_stats <- bind_rows(
  
  data2015_clean %>%
    transmute(
      year = 2015,
      Senior = senior_la_pct,
      SNAP = snap_la_pct
    ),
  
  data2019_clean %>%
    transmute(
      year = 2019,
      Senior = senior_la_pct,
      SNAP = snap_la_pct
    )
  
) %>%
  
  pivot_longer(
    cols = c(Senior, SNAP),
    names_to = "Group",
    values_to = "Low_Access_Pct"
  ) %>%
  
  group_by(year, Group) %>%
  
  summarise(
    Min = min(Low_Access_Pct, na.rm = TRUE),
    Q1 = quantile(Low_Access_Pct, 0.25, na.rm = TRUE),
    Median = median(Low_Access_Pct, na.rm = TRUE),
    Mean = mean(Low_Access_Pct, na.rm = TRUE),
    Q3 = quantile(Low_Access_Pct, 0.75, na.rm = TRUE),
    Max = max(Low_Access_Pct, na.rm = TRUE),
    Missing = sum(is.na(Low_Access_Pct)),
    .groups = "drop"
  ) %>%
  
  mutate(
    across(
      c(Min, Q1, Median, Mean, Q3, Max),
      ~ round(.x, 2)
    )
  )

summary_stats
```

```{r, echo = FALSE, include=FALSE}
kable(
  summary_stats,
  caption = "Summary Statistics for Senior and SNAP Low-Access Shares, 2015 and 2019",
  col.names = c(
    "Year", "Group", "Min (%)", "Q1 (%)", "Median (%)", "Mean (%)", "Q3 (%)", "Max (%)", "Missing"
  ),
  align = c(
    "c", "l",
    "r", "r", "r", "r", "r", "r", "r"
  )
)
```

```{r, echo = FALSE, include=FALSE}
# Check whether over 100
data2015_clean %>%
  summarise(
    senior_over_100 = sum(senior_la_pct > 100, na.rm = TRUE),
    snap_over_100 = sum(snap_la_pct > 100, na.rm = TRUE)
  )

data2019_clean %>%
  summarise(
    senior_over_100 = sum(senior_la_pct > 100, na.rm = TRUE),
    snap_over_100 = sum(snap_la_pct > 100, na.rm = TRUE)
  )
```

```{r, echo = FALSE, include=FALSE}
data2015_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )

data2019_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )
```

```{r, echo = FALSE, include=FALSE}
options(digits = 15)

data2015_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )

data2019_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )
```

```{r, echo = FALSE, include=FALSE}
combined_data <- bind_rows(
  data2015_clean,
  data2019_clean
)
combined_data
```

```{r, echo = FALSE, include=FALSE}
dim(combined_data)
table(combined_data$year)
table(combined_data$area, useNA = "ifany")
```

After harmonisation, the 2015 and 2019 datasets were combined into a
dataset containing 145,395 census-tract observations. Of these, 110,341
were classified as urban and 35,054 as rural, with no missing values in
the urban–rural classification.

```{r, echo = FALSE, include=FALSE}
summary_table <- combined_data %>%
  group_by(year, area) %>%
  summarise(
    n_tracts = n(),

    n_valid_senior =
      sum(!is.na(senior_la_pct)),

    n_valid_snap =
      sum(!is.na(snap_la_pct)),

    mean_senior_pct =
      mean(senior_la_pct, na.rm = TRUE),

    median_senior_pct =
      median(senior_la_pct, na.rm = TRUE),

    mean_snap_pct =
      mean(snap_la_pct, na.rm = TRUE),

    median_snap_pct =
      median(snap_la_pct, na.rm = TRUE),

    .groups = "drop"
  )

summary_table
```

```{r, echo = FALSE, include=FALSE}
summary_table_report <- summary_table %>%
  mutate(
    across(
      c(
        mean_senior_pct,
        median_senior_pct,
        mean_snap_pct,
        median_snap_pct
      ),
      ~ round(.x, 2)
    )
  )

summary_table_report
```

```{r, echo = FALSE, include=FALSE}
dim(data2015_clean)
names(data2015_clean)
str(data2015_clean)
summary(data2015_clean)
```

```{r, echo = FALSE, include=FALSE}
dim(data2019_clean)
names(data2019_clean)
str(data2019_clean)
summary(data2019_clean)
```

```{r, echo = FALSE, include=FALSE}
colSums(is.na(data2019_clean))
```

```{r, echo = FALSE, include=FALSE}
colSums(is.na(data2015_clean))
```

\newpage


# Exploratory Data Analysis

**3.1 Overview of the Analysis**

The exploratory analysis focused on the low-supermarket-access shares
for two groups: adults aged 65 years and over, and SNAP-receiving households.
Comparisons were made between 2015 and 2019 and between urban and rural
census tracts. For urban census tracts, low supermarket access was
measured using the one-mile threshold, while the ten-mile threshold was
used for rural census tracts. All low-access shares were expressed as
percentages to ensure consistency between the 2015 and 2019 datasets.Because 
urban and rural low-accessibility percentages are not calculated on the same 
scale, the report's urban-rural comparisons reflect each area's results under 
USDA standards—not a direct distance comparison. This analysis covers 2015–2019 
changes and urban-rural tract differences, with all figures in percentages for
consistent year-over-year comparison.

Table 1 shows a clear unbran-rural gap: the mean low-access share for 
older adults in urban census tracts is roughly double that in rural tracts 
(3.4% vs. 1.7% in 2015; 5.3% vs. 4.0% in 2019), and the same pattern holds for 
SNAP households (1.3% vs. 2.9% in 2015; 2.8% vs. 4.3% in 2019). This pattern 
reflects the differing USDA thresholds applied to urban and rural areas, rather 
than an absolute access gap. The effective sample size dropped substantially 
between 2015 and 2019, with rural tract observations more than halving 
(from 17,692 to around 7,400), suggesting that 2019 estimates should 
be interpreted with greater caution than those for 2015. Based on the table, 
every mean exceeds its corresponding median, with the largest gap observed 
in the 2015 rural group (mean 1.7% vs. median 0.0%), indicating a strongly 
right-skewed distribution where a small number of tracts with particularly 
severe accessibility barriers pull the averages upward.

```{r, echo = FALSE}
summary_table <- combined_data %>%
  group_by(area, year) %>%
  summarise(
    Senior_Mean   = round(mean(senior_la_pct, na.rm = TRUE), 1),
    Senior_Median = round(median(senior_la_pct, na.rm = TRUE), 1),
    Senior_Valid_N = sum(!is.na(senior_la_pct)),
    
    SNAP_Mean     = round(mean(snap_la_pct, na.rm = TRUE), 1),
    SNAP_Median   = round(median(snap_la_pct, na.rm = TRUE), 1),
    SNAP_Valid_N = sum(!is.na(snap_la_pct)),
    n_tracts      = n(),
    .groups = "drop"
  )

kable(summary_table,
      caption = "Mean and median low-access share (%) for seniors and SNAP households, by area and year.",
      col.names = c("Area", "Year", "Senior Mean", "Senior Median", "Senior N (Valid)",
                    "SNAP Mean", "SNAP Median", "SNAP N (Valid)", "N Tracts"))
```

```{r, echo = FALSE, include = FALSE}
missing_table <- data.frame(
  Year = c(2015, 2019),
  Senior_Missing = c(
    sum(is.na(data2015_clean$senior_la_pct)),
    sum(is.na(data2019_clean$senior_la_pct))
  ),
  SNAP_Missing = c(
    sum(is.na(data2015_clean$snap_la_pct)),
    sum(is.na(data2019_clean$snap_la_pct))
  )
)

kable(missing_table,
      caption = "Number of missing Senior and SNAP low-access share observations, by year.",
      col.names = c("Year", "Senior Missing", "SNAP Missing"))
```


**3.2 Senior Low-Access Share: Distribution and Urban–Rural Comparison**

The following sections look at distribution of variables in more detail, 
starting with seniors.

Figure 1 shows the distribution of senior low-access shares by year and
urban/rural areas. The distributions were strongly right-skewed across
all four groups, with most census tracts having relatively low senior
low-access shares and a smaller number having much higher values. Urban
census tracts generally showed higher low-access shares than rural
census tracts. The 2019 distributions also appeared shifted toward
higher values than in 2015. However, interpret the 2019 results
cautiously because the senior low-access variable contained substantial
missing data.

```{r, echo = FALSE,fig.width = 5.8, fig.height = 3.0}
ggplot(combined_data, aes(x = senior_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  facet_grid(area ~ year) +
  labs(
    title = "Figure1. Distribution of Senior Low-Access Share",
    x = "Senior Low-Access Share (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Figure 2 provides a more direct comparison between urban and rural tracts 
for the senior low-access share. In both 2015 and 2019, the median senior 
low-access share was consistently higher in urban tracts than in rural tracts, 
rising from 0.6% to 3.5% in urban areas compared with 0.0% to 1.7% in rural 
areas. The spread also widened between the two years: in rural areas, the 
boxp shifted from being almost flat at zero in 2015, but by 2019 it had 
extended noticeably upward by several percentage points. Urban tracts showed 
a similar pattern, with the box stretching further above the median, 
indicating greater variability across tracts in 2019. In both years, a number 
of outlying high values are visible, which is consistent with the right-skewed 
distribution observed in Figure 1.

```{r, echo = FALSE,warning=FALSE, fig.width = 5.8, fig.height = 3.0}
# Senior Boxplot
ggplot(combined_data, 
       aes(x = factor(year), y = senior_la_pct, fill = area)) +
  geom_boxplot(outlier.size = 0.3,na.rm = TRUE) +
  scale_fill_grey() +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = sprintf("%.1f", after_stat(y))),
    position = position_dodge(width = 0.75),
    vjust = -0.8,
    size = 3,
  ) +
  labs(
    title = "Figure 2. Senior Low-Access Share by Year and Area",
    x = "Year",
    y = "Senior Low-Access Share (%)",
    fill = "Area"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Overall, the senior low-access share was generally higher in urban than rural 
census tracts, and the distributions were strongly right-skewed in both years. 
The apparent increase from 2015 to 2019 may be partly attributable to 
differences in the set of census tracts with available data, rather than 
reflecting a genuine change. In 2019, values for low accessibility among older 
adults were missing for approximately 30,000 census tracts. This issue is 
addressed directly in the matched analysis presented in Section 3.4.

**3.3 SNAP Low-Access Share: Distribution and Urban–Rural Comparison**

This section examines the SNAP low-access share in more detail.
```{r, echo = FALSE,include = FALSE}
ggplot(combined_data, aes(x = snap_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(
    title = "Figure 3: Distribution of SNAP Low-Access Share, All Years and Areas",
    x = "SNAP Low-Access Share (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

Figure 3 displays the distribution of SNAP low-access share by year and 
urban-rural regions. The distributions of all four groups exhibit a strong 
right-skewed pattern: the SNAP low-access share in most census areas are 
relatively low, while only a few census areas have significantly higher values. 
Urban census areas show a longer and more extended tail to the right in both 
years compared to rural census areas, indicating greater variability among 
urban census areas.

```{r, echo = FALSE, fig.width = 5.8, fig.height = 3.0}
ggplot(combined_data,aes(x = snap_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  facet_grid(area ~ year) +
  labs(
    title = "Figure 3. Distribution of SNAP Low-Access Share",
    x = "SNAP Low-Access Share (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Figure 4 shows that the median SNAP low-access share was consistently higher in 
urban tracts than in rural tracts in both years, rising from 0.3% to 1.8% in 
urban areas compared with 0.0% to 1.0% in rural areas. The spread also widened 
in both areas: rural boxes expanded from nearly flush with zero to about one 
percentage point higher, while urban boxes stretched further above the median, 
indicating greater tract-level variation in 2019. The distribution in both 
years and for both regions remained strongly skewed to the right, with high 
values observed extending to approximately 85% - 100%, indicating that a small 
portion of the surveyed areas were more pronounced in urban areas and continued 
to face significantly higher barriers to supermarket accessibility compared to 
typical surveyed areas.

```{r, echo = FALSE, warning=FALSE,fig.width = 5.8, fig.height = 3.0}
# SNAP boxplot
ggplot(combined_data, 
       aes(x = factor(year), y = snap_la_pct,fill = area)) +
  geom_boxplot(outlier.size = 0.3, na.rm = TRUE) +
  scale_fill_grey() +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = sprintf("%.1f", after_stat(y))),
    position = position_dodge(width = 0.75),
    vjust = -0.5,
    size = 3
  ) +
  labs(
    title = "Figure 4. SNAP Low-Access Share by Year and Area",
    x = "Year",
    y = "SNAP Low-Access Share (%)",
    fill = "Area"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

In short, the SNAP low-access share was generally higher in urban than rural 
census tracts, and the distributions were strongly right-skewed in both years. 
The apparent increase from 2015 to 2019 may be partly attributable to 
differences in the set of census tracts with available data, rather than 
reflecting a genuine change. In 2019, values for SNAP low-access were missing 
for approximately 29,918 census tracts. This issue is addressed directly in 
the matched analysis presented in Section 3.4.

**3.4 Changes between 2015 and 2019**

To examine changes between 2015 and 2019 more directly, census tracts in
the two datasets were matched using their Census Tract identifiers. A total 
of 72,506 census tracts were successfully matched across the two years. Of 
these, 21 census tracts changed their urban/rural classification between 2015 
and 2019. These tracts were excluded from the change analysis because the 
low-access distance threshold differs between urban and rural areas. This 
resulted in 72,485 census tracts with a stable urban/rural classification.

```{r, echo = FALSE, include = FALSE}
# Match the same census tract in 2015 and 2019, and calculate the changes in the share of seniors and SNAP low-access individuals between the two years.

matched_data <- data2015_clean %>%
  select(
    censustract,
    area_2015 = area,
    senior_pct_2015 = senior_la_pct,
    snap_pct_2015 = snap_la_pct
  ) %>%
  inner_join(
    data2019_clean %>%
      select(
        censustract,
        area_2019 = area,
        senior_pct_2019 = senior_la_pct,
        snap_pct_2019 = snap_la_pct
      ),
    by = "censustract"
  ) %>%
  mutate(
    senior_change = senior_pct_2019 - senior_pct_2015,
    snap_change = snap_pct_2019 - snap_pct_2015
  )

matched_data
```

```{r, echo = FALSE, include = FALSE}
dim(matched_data)
sum(!is.na(matched_data$senior_change))
sum(!is.na(matched_data$snap_change))
```

```{r, echo = FALSE, include = FALSE}
# Senior Change (using all matched tracts, before excluding reclassified tracts)
ggplot(
  matched_data,
  aes(x = area_2015, y = senior_change)) +
  geom_boxplot(na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Change in Senior Low-Access Share, 2015-2019",
    x = "Area",
    y = "Change in percentage points"
  ) +
  theme_minimal()
```

```{r, echo = FALSE, include = FALSE}
# Check the boxplot why is the line 
matched_data %>%
  group_by(area_2015) %>%
  summarise(n = n(), 
            Q1 = quantile(senior_change, 0.25, na.rm = TRUE),
            Median = median(senior_change, na.rm = TRUE),
            Q3 = quantile(senior_change, 0.75, na.rm = TRUE),
            Min = min(senior_change, na.rm = TRUE),
            Max = max(senior_change, na.rm = TRUE),
            pct_exact_zero =mean(senior_change == 0,na.rm = TRUE) * 100
  )

```

```{r, echo = FALSE, include = FALSE}
matched_data %>%
  group_by(area_2015) %>%
  summarise(
    pct_near_zero =
      mean(
        abs(senior_change) < 1e-8,
        na.rm = TRUE
      ) * 100
  )
```

```{r, echo = FALSE, include = FALSE}
# SNAP
ggplot(
  matched_data,
  aes(x = area_2015, y = snap_change)) +
  geom_boxplot(na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Change in SNAP Low-Access Share, 2015-2019",
    x = "Area",
    y = "Change in percentage points"
  ) +
  theme_minimal()
```

```{r, echo = FALSE, include = FALSE}
matched_stable <- matched_data %>%
  filter(area_2015 == area_2019)

matched_stable
```

For the paired analysis, only census tracts with non-missing low-access
shares in both 2015 and 2019 were included. Of the 72,485 tracts with stable 
classification, 42,554 had complete senior data in both years and 42,589 had 
complete SNAP data in both years (Table 2). Change was calculated as:
Change= Low-access share2019 - Low-access share2015 and is reported in
percentage points. Positive values therefore represent an increase in
low-access share between 2015 and 2019, while negative values represent
a decrease.

```{r, echo = FALSE, include = FALSE}
change_summary <- matched_data %>%
  group_by(area_2015) %>%
  summarise(
    n = sum(!is.na(senior_change)),
    median_change =median(senior_change, na.rm = TRUE),
    mean_change =mean(senior_change, na.rm = TRUE),
    pct_increase =mean(senior_change > 0.1, na.rm = TRUE) * 100,
    pct_decrease =mean(senior_change < -0.1, na.rm = TRUE) * 100,
    pct_stable = mean(abs(senior_change) <= 0.1, na.rm = TRUE) * 100,
    .groups = "drop"
  )

change_summary
```

```{r, echo = FALSE, include = FALSE}
table(
  matched_data$area_2015,
  matched_data$area_2019
)
```

```{r, echo = FALSE, include = FALSE}
# Make sure the number of row between two datasets
nrow(matched_data)
nrow(matched_stable)

nrow(matched_data) - nrow(matched_stable)
```

```{r, echo = FALSE, include = FALSE}
matched_stable <- matched_stable %>%
  mutate(snap_change = snap_pct_2019 - snap_pct_2015)

matched_stable
```

```{r, echo = FALSE, include = FALSE}
senior_paired_data <- matched_stable %>%
  filter(
    !is.na(senior_pct_2015),
    !is.na(senior_pct_2019)
  )

senior_paired <- senior_paired_data %>%
  group_by(area_2015) %>%
  summarise(
    N = n(),
    
    Mean_2015 = mean(senior_pct_2015),
    Mean_2019 = mean(senior_pct_2019),
    
    Median_2015 = median(senior_pct_2015),
    Median_2019 = median(senior_pct_2019),
    
    Mean_Change = mean(senior_change),
    Median_Change = median(senior_change),
    
    .groups = "drop"
  ) %>%
  mutate(
    across(
      c(
        Mean_2015,
        Mean_2019,
        Median_2015,
        Median_2019,
        Mean_Change,
        Median_Change
      ),
      ~ round(.x, 2)
    )
  )

kable(
  senior_paired,
  caption = "Paired summary of Senior low-access share, 2015-2019",
  col.names = c(
    "Area",
    "N",
    "Mean 2015 (%)",
    "Mean 2019 (%)",
    "Median 2015 (%)",
    "Median 2019 (%)",
    "Mean Change (pp)",
    "Median Change (pp)"
  )
)
```

```{r, echo = FALSE, include = FALSE}
snap_paired_data <- matched_stable %>%
  filter(
    !is.na(snap_pct_2015),
    !is.na(snap_pct_2019)
  )

snap_paired <- snap_paired_data %>%
  group_by(area_2015) %>%
  summarise(
    N = n(),
    
    Mean_2015 = mean(snap_pct_2015),
    Mean_2019 = mean(snap_pct_2019),
    
    Median_2015 = median(snap_pct_2015),
    Median_2019 = median(snap_pct_2019),
    
    Mean_Change = mean(snap_change),
    Median_Change = median(snap_change),
    
    .groups = "drop"
  ) %>%
  mutate(
    across(
      c(
        Mean_2015,
        Mean_2019,
        Median_2015,
        Median_2019,
        Mean_Change,
        Median_Change
      ),
      ~ round(.x, 2)
    )
  )

kable(
  snap_paired,
  caption = "Paired summary of SNAP low-access share, 2015-2019",
  col.names = c(
    "Area",
    "N",
    "Mean 2015 (%)",
    "Mean 2019 (%)",
    "Median 2015 (%)",
    "Median 2019 (%)",
    "Mean Change (pp)",
    "Median Change (pp)"
  )
)
```

Table 2 reveals an interesting contrast: the trends for senior and SNAP 
households moved in opposite directions between 2015 and 2019. The mean 
proportion of older adults rose slightly in both rural (from 3.94% to 3.96%) 
and urban (from 5.21% to 5.32%) census tracts, whilst the mean proportion of 
SNAP recipients fell slightly in both rural (from 2.91% to 2.77%) and urban 
(from 4.40% to 4.33%) census tracts. These opposite trends suggest that 
changes in supermarket accessibility for these two groups are not driven by 
the same underlying trend, indicating that the factors influencing changes 
in accessibility for older people may not be entirely the same as those 
affecting SNAP households. However, the magnitude of these changes is small 
relative to the degree of variation within each group shown in Table 1, and 
the median changes for all four groups are exactly zero.

```{r, echo = FALSE}
paired_summary <- bind_rows(
  senior_paired %>%
    mutate(Group = "Senior"),
  
  snap_paired %>%
    mutate(Group = "SNAP")
) %>%
  select(
    Group,
    area_2015,
    N,
    Mean_2015,
    Mean_2019,
    Median_2015,
    Median_2019,
    Mean_Change,
    Median_Change
  ) %>%
  mutate(across(c(Mean_2015, Mean_2019, Median_2015, Median_2019,
                Mean_Change, Median_Change),
              ~ {
                x <- round(.x, 2)
                x[x == 0] <- 0
                sprintf("%.2f", x)
              })) %>%
  arrange(Group, area_2015)

kable(
  paired_summary,
  caption = "Paired summary of Senior and SNAP low-access shares, 2015-2019",
  col.names = c(
    "Group",
    "Area",
    "N",
    "Mean 2015 (%)",
    "Mean 2019 (%)",
    "Median 2015 (%)",
    "Median 2019 (%)",
    "Mean Change (pp)",
    "Median Change (pp)"
  ),  
  align = c(
    "c", "l",
    "r", "r", "r", "r", "r", "r", "r"
  )
)
```

```{r, echo=FALSE, include=FALSE}
#matched_stable %>%
#  group_by(area_2015) %>%
#  summarise(
#    n = sum(!is.na(snap_change)),
#    Q1 = quantile(snap_change, 0.25, na.rm = TRUE),
#    Median = median(snap_change, na.rm = TRUE),
#    Q3 = quantile(snap_change, 0.75, na.rm = TRUE),
#    Min = min(snap_change, na.rm = TRUE),
#    Max = max(snap_change, na.rm = TRUE),
#    .groups = "drop"
#  )

#matched_stable %>%
#  group_by(area_2019) %>%
#  summarise(
#    n = sum(!is.na(snap_change)),
#    Q1 = quantile(snap_change, 0.25, na.rm = TRUE),
#    Median = median(snap_change, na.rm = TRUE),
#    Q3 = quantile(snap_change, 0.75, na.rm = TRUE),
#    Min = min(snap_change, na.rm = TRUE),
#    Max = max(snap_change, na.rm = TRUE),
#    .groups = "drop"
#  )

snap_change_summary <- matched_stable %>%
  group_by(Area = area_2015) %>%
  summarise(
    N = sum(!is.na(snap_change)),
    Q1 = round(quantile(snap_change, 0.25, na.rm = TRUE), 2),
    Median = round(median(snap_change, na.rm = TRUE), 2),
    Q3 = round(quantile(snap_change, 0.75, na.rm = TRUE), 2),
    Min = round(min(snap_change, na.rm = TRUE), 2),
    Max = round(max(snap_change, na.rm = TRUE), 2),
    .groups = "drop"
  )

kable(
  snap_change_summary,
  caption = "Summary of changes in SNAP low-access share, 2015–2019, by area"
)
```

```{r,echo=FALSE,include=FALSE, fig.width = 5.8, fig.height = 3.0}
ggplot(matched_data,aes(x = senior_change)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  facet_wrap(~ area_2015) +
  geom_vline(xintercept = 0,linetype = "dashed") +
  labs(
    title = "Distribution of Changes in Senior Low-Access Share, 2015-2019 (unfiltered)",
    x = "Change in percentage points",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Figure 5 shows that changes in senior low-access share were strongly
concentrated around zero in both rural and urban census tracts. This is
consistent with the zero median changes reported in Table 2. Although
most census tracts experienced relatively small changes, a smaller
number showed substantially larger positive or negative changes.

```{r,echo=FALSE, fig.width = 5.8, fig.height = 3.0, out.width="80%"}
ggplot(matched_stable, aes(x = senior_change)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  facet_wrap(~ area_2015) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Figure 5. Distribution of Changes in Senior Low-Access Share (2015-2019)",
    x = "Change in percentage points",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Figure 6 shows that change distributions in SNAP Low-Access share were also 
strongly concentrated around zero, although some census tracts experienced 
larger increases or decreases. This is consistent with the zero median changes 
reported in Table 2, and together they indicate that the overall changes 
between the two years were small for most matched census tracts.

```{r,echo=FALSE, fig.width = 5.8, fig.height = 3.0, out.width="80%"}
ggplot(matched_stable,aes(x = snap_change)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  facet_wrap(~ area_2015) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Figure 6. Distribution of Changes in SNAP Low-Access Share (2015-2019)",
    x = "Change in percentage points",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

```{r,echo=FALSE,include=FALSE}
matched_stable %>%
  group_by(area_2015) %>%
  summarise(
    n = sum(!is.na(snap_change)),
    Q1 = quantile(snap_change,0.25,na.rm = TRUE),
    Median = median(snap_change,na.rm = TRUE),
    Q3 = quantile(snap_change,0.75,na.rm = TRUE),
    Min = min(snap_change,na.rm = TRUE),
    Max = max(snap_change,na.rm = TRUE),
    pct_near_zero =mean(abs(snap_change) < 1e-8,na.rm = TRUE) * 100,
    .groups = "drop"
  )
```

In summary, the paired analysis illustrates that the changes at the census 
tract level between 2015 and 2019 were generally small. Senior and SNAP 
low-access shares moved in opposite directions on average, but median changes 
were zero across all four groups, indicating that the small mean shifts were 
driven by a minority of tracts with larger changes rather than a broad trend. 
These findings are descriptive results and do not establish statistical 
significance; the paired analysis itself cannot eliminate the potential bias 
caused by missing data. If there are systematic differences between the 
census tracts with missing 2019 data and those with complete data, the above 
paired results may still be affected.

**3.5 Additional Comparison: Mean Low-Access Share by Group**

To talk about... median much lower then means = skewed data. Urban
vastly outnumber rural. Values increase across all four rows

keeping in mind a rural tract requires households and seniors to be at
least 10 miles away compared to a min of 1

```{r, echo=FALSE, fig.width=5.8, fig.height=3.0, out.width="90%"}
library(tidyr)

dot_data <- summary_table %>%
  select(area, year, Senior_Mean, SNAP_Mean) %>%
  pivot_longer(cols = c(Senior_Mean, SNAP_Mean),
               names_to = "Group", values_to = "Mean_Rate") %>%
  mutate(Group = recode(Group, Senior_Mean = "Seniors", SNAP_Mean = "SNAP"))

ggplot(dot_data, aes(x = Mean_Rate, y = Group, color = factor(year))) +
  geom_point(size = 3) +
  facet_wrap(~area) +
  labs(
    title = "Mean Low-Access Share by Group, Area, and Year",
    x = "Mean Low-Access Share (%)",
    y = NULL,
    color = "Year"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Across both Senior and SNAP low-access shares, 2019 shows a higher mean
rate than 2015, and Urban tracts show a higher mean rate than Rural
tracts, in both years and for both groups

Because "low-access" is defined using a much stricter distance threshold
in urban areas (1 mile) than rural areas (10 miles), a higher urban
low-access rate does not necessarily mean urban residents travel further
to reach a supermarket than rural residents. This may just reflect that
urban areas are held to a stricter access standard. A rural resident
traveling several miles to a store may still count as "having access"
under the rural threshold, while a much shorter distance in a city would
count as "low access."

As a result, the lower mean low-access share in rural areas should not
be read as evidence that rural food access is genuinely better. Rural
residents may need to travel considerably farther in absolute terms
before being classified as low-access at all, meaning the rural rate
likely understates the true burden of distance compared to the urban
rate. This distinction should be kept in mind whenever urban and rural
low-access rates are compared directly in this report, rather than
treating them as measuring the same underlying concept.

**3.6 Summary**

In conclusion, the exploratory analysis has identified several important 
patterns in the indicators of senior and SNAP low-access. First, both 
low-access share distributions were strongly right-skewed. Most census 
tracts had relatively low low-access shares, while a smaller number of 
tracts had substantially higher values. Thus, the mean was generally 
higher than the median, indicating that the mean was influenced by census
tracts with relatively large low-access shares.

Across both measures, urban census tracts consistently showed higher low-access 
shares than rural tracts in both years. However, since the urban census tracts 
use a stricter one-mile threshold while the rural census tracts use a ten-mile 
threshold, this disparity reflects the classification results of each region 
based on their own USDA standards, rather than directly measuring which group 
of residents is actually farther from the supermarket.

Although cross-sectional distributions in 2019 appeared to be higher than
those in 2015, such comparisons should be interpreted with caution as there 
were a large number of missing values in the 2019 data.The paired analysis 
provided a more direct comparison of the same census tracts across years and 
showed that changes were generally small. Senior low-access shares increased 
slightly on average, while SNAP low-access shares decreased slightly, and the 
median change was zero across both rural and urban census tracts.


\newpage

# Individual Contributions




