---
title: "FoodAccessPREP"
author: "Oliver Temel"
date: "2026-08-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Background and Data**

Our group used the 2015 and 2019 Food Access Research Atlas datasets, available through the U.S. Department of Agriculture Economic Research Service.

The datasets are of interest because they provide insight into access to supermarkets and essential food resources across different communities. They can be used to examine food access among people receiving benefits, those living in areas of high deprivation or low socioeconomic status, those living further away in urban vs rural areas, and different demographic groups. Using data from multiple years also allows us to identify changes over time and determine whether there have been noticeable improvements in food access.

The datasets contain information including census tract, poverty rate, food access at specified distances, and access measures across different racial and age groups, as well as SNAP beneficiaries. The majority of the variables are numerical, with some categorical variables such as census tract, state, and county. The data is structured so that individual variables represent specific measures of food access for particular demographic groups and distance thresholds.

One point of interest was that the 2019 dataset contained a large number of NULL values. These were assumed to represent zero values, as many of the affected variables represented whether an observation belonged to a particular demographic or access category. For example, a variable could represent the number of Native American individuals experiencing low food access at 20 miles, where a NULL indicated that there were no individuals meeting that condition. In contrast, both datasets also contained a considerable number of genuine NA values, particularly for SNAP_la_rate (2,405) and senior_la_rate (520).

To enable comparisons between years, our group used both the 2015 and 2019 Food Access Research Atlas datasets. The datasets were straightforward to integrate because they contained consistent and largely homogeneous variable names. We therefore retained the same variables from each dataset, allowing us to make meaningful comparisons in food access between 2015 and 2019.

# DATA 2015

```{r Data Inputs 2015}
library(readxl)
DATA2015 <- read_excel("FoodAccessResearchAtlasData2015.xlsx", sheet = 3)
head(DATA2015)
```

# DATA 2019

```{r DATA Inputs 2019}
DATA2019 <- read_excel("FoodAccessResearchAtlasData2019.xlsx", sheet = 3)
head(DATA2019)
```

```{r}
dim(DATA2015)
dim(DATA2019)

names(DATA2015)
names(DATA2019)

str(DATA2015)
str(DATA2019)

```

```{r}
names(DATA2015) <- tolower(names(DATA2015))
names(DATA2019) <- tolower(names(DATA2019))
```


```{r}
library(dplyr)
data2015_clean <- DATA2015 %>%
  select(
    censustract,
    state,
    county,
    urban,
    laseniors1,
    laseniors10,
    lasnap1,
    lasnap10,
    tractseniors,
    tractsnap,
    laseniors20,
    laseniorshalf,
    laseniorshalfshare
  )

data2019_clean <- DATA2019 %>%
  select(
    censustract,
    state,
    county,
    urban,
    laseniors1,
    laseniors10,
    lasnap1,
    lasnap10,
    tractseniors,
    tractsnap,
    laseniors20,
    laseniorshalf,
    laseniorshalfshare
  )
```

```{r add year}
data2015_clean <- data2015_clean %>%
  mutate(year = 2015)

data2019_clean <- data2019_clean %>%
  mutate(year = 2019)
```

```{r}
#Create Urban/Rural
data2015_clean <- data2015_clean %>%
  mutate(
    area = if_else(
      urban == 1,
      "Urban",
      "Rural"
    )
  )

data2019_clean <- data2019_clean %>%
  mutate(
    area = if_else(
      urban == 1,
      "Urban",
      "Rural"
    )
  )
```


```{r}
# Seniors
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_n = if_else(
      urban == 1,
      laseniors1,
      laseniors10
    )
  )

# SNAP
data2015_clean <- data2015_clean %>%
  mutate(
    snap_la_n = if_else(
      urban == 1,
      lasnap1,
      lasnap10
    )
  )

# Seniors
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_n = if_else(
      urban == 1,
      laseniors1,
      laseniors10
    )
  )

# SNAP
data2019_clean <- data2019_clean %>%
  mutate(
    snap_la_n = if_else(
      urban == 1,
      lasnap1,
      lasnap10
    )
  )
```

```{r}
summary(data2015_clean)
summary(data2019_clean)
```

```{r}
data2019_clean <- data2019_clean %>%
  mutate(
    across(
      c(laseniors1, laseniors10, lasnap1, lasnap10,
        laseniors20,laseniorshalf,laseniorshalfshare,senior_la_n, snap_la_n),
      ~ as.numeric(ifelse(.x == "NULL", "0", .x))
    )
  )
```

```{r}
summary(data2019_clean)
```


```{r}
#Senior
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_rate = if_else(
      tractseniors > 0,
      senior_la_n / tractseniors,
      NA_real_
    )
  )

# SNAP
data2015_clean <- data2015_clean %>%
  mutate(
    snap_la_rate = if_else(
      tractsnap > 0,
      snap_la_n / tractsnap,
      NA_real_
    )
  )

#Senior
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_rate = if_else(
      tractseniors > 0,
      senior_la_n / tractseniors,
      NA_real_
    )
  )

# SNAP
data2019_clean <- data2019_clean %>%
  mutate(
    snap_la_rate = if_else(
      tractsnap > 0,
      snap_la_n / tractsnap,
      NA_real_
    )
  )
```

```{r}
# Show precentage
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_pct = senior_la_rate * 100,
    snap_la_pct = snap_la_rate * 100
  )

data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_pct = senior_la_rate * 100,
    snap_la_pct = snap_la_rate * 100
  )
```

```{r}
sapply(
  data2015_clean[
    c(
      "urban",
      "laseniors1",
      "laseniors10",
      "lasnap1",
      "lasnap10",
      "tractseniors",
      "tractsnap",
      "laseniors20",
      "laseniorshalf",
      "laseniorshalfshare"
    )
  ],
  class
)

sapply(
  data2019_clean[
    c(
      "urban",
      "laseniors1",
      "laseniors10",
      "lasnap1",
      "lasnap10",
      "tractseniors",
      "tractsnap",
      "laseniors20",
      "laseniorshalf",
      "laseniorshalfshare"
    )
  ],
  class
)
```


```{r}
numeric_vars <- c(
  "urban",
  "laseniors1",
  "laseniors10",
  "lasnap1",
  "lasnap10",
  "tractseniors",
  "tractsnap",
  "senior_la_n",
  "snap_la_n",
  "senior_la_rate",
  "snap_la_rate",
  "senior_la_pct",
  "snap_la_pct",
  "laseniors20",
  "laseniorshalf",
  "laseniorshalfshare"
)

data2015_clean <- data2015_clean %>%
  mutate(
    across(
      any_of(numeric_vars),
      ~ as.numeric(.)
    )
  )

data2019_clean <- data2019_clean %>%
  mutate(
    across(
      any_of(numeric_vars),
      ~ as.numeric(.)
    )
  )

combined_data <- bind_rows(
  data2015_clean,
  data2019_clean
)

combined_data
```

```{r}
summary_table <- combined_data %>%
  group_by(
    year,
    area
  ) %>%

  summarise(

    n_tracts = n(),

    senior_la_rate =
      sum(senior_la_n, na.rm = TRUE) /
      sum(tractseniors, na.rm = TRUE),

    snap_la_rate =
      sum(snap_la_n, na.rm = TRUE) /
      sum(tractsnap, na.rm = TRUE),

    .groups = "drop"
  ) %>%

  mutate(
    senior_la_pct =
      senior_la_rate * 100,

    snap_la_pct =
      snap_la_rate * 100
  )

summary_table
```
```{r}
dim(data2015_clean)
names(data2015_clean)
str(data2015_clean)
summary(data2015_clean)
```
Data Cleaning and Missing Data. 

```{r}
dim(data2019_clean)
names(data2019_clean)
str(data2019_clean)
summary(data2019_clean)
```

```{r}
colSums(is.na(data2019_clean))

```

```{r}
dim(data2015_clean)
names(data2015_clean)
str(data2015_clean)
summary(data2015_clean)
```

```{r}
colSums(is.na(data2015_clean))
```

```{r}
summary(data2015_clean$snap_la_rate)
summary(data2019_clean$snap_la_rate)
```

```{r}
summary(data2015_clean$senior_la_rate)
summary(data2019_clean$senior_la_rate)
```

```{r}
library(ggplot2)
ggplot(data2019_clean, aes(x = snap_la_rate)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(
    title = "Distribution of SNAP Low-Access Rate, 2019",
    x = "SNAP Low-Access Rate",
    y = "Number of Census Tracts"
  )
```

```{r}
library(ggplot2)
ggplot(data2015_clean, aes(x = snap_la_rate)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(
    title = "Distribution of SNAP Low-Access Rate, 2015",
    x = "SNAP Low-Access Rate",
    y = "Number of Census Tracts"
  )
```
```{r}
library(ggplot2)
ggplot(data2019_clean, aes(x = senior_la_rate)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(
    title = "Distribution of Senior Low-Access Rate, 2019",
    x = "Senior Low-Access Rate",
    y = "Number of Census Tracts"
  )
```

```{r}
library(ggplot2)
ggplot(data2015_clean, aes(x = senior_la_rate)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(
    title = "Distribution of Senior Low-Access Rate, 2015",
    x = "Senior Low-Access Rate",
    y = "Number of Census Tracts"
  )
```