---
title: "FoodAccessPREP"
author: "Oliver Temel, Mairangi Campbell and Jiayi Lu"
date: "2026-08-19"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background and Data

Our group used the 2015 and 2019 Food Access Research Atlas datasets,
available through the U.S. Department of Agriculture Economic Research
Service.

The datasets are of interest because they provide insight into access to
supermarkets and essential food resources across different communities.
They can be used to examine food access among people receiving benefits,
those living in areas of high deprivation or low socioeconomic status,
those living further away in urban vs rural areas, and different
demographic groups. Using data from multiple years also allows us to
identify changes over time and determine whether there have been
noticeable improvements in food access.

The datasets contain information including census tract, poverty rate,
food access at specified distances, and access measures across different
racial and age groups, as well as SNAP beneficiaries. The majority of
the variables are numerical, with some categorical variables such as
census tract, state, and county. The data is structured so that
individual variables represent specific measures of food access for
particular demographic groups and distance thresholds.

One point of interest was that the 2019 dataset contained a large number
of NULL values. These were assumed to represent zero values, as many of
the affected variables represented whether an observation belonged to a
particular demographic or access category. For example, a variable could
represent the number of Native American individuals experiencing low
food access at 20 miles, where a NULL indicated that there were no
individuals meeting that condition. In contrast, both datasets also
contained a considerable number of genuine NA values, particularly for
SNAP_la_rate (2,405) and senior_la_rate (520).

To enable comparisons between years, our group used both the 2015 and
2019 Food Access Research Atlas datasets. The datasets were
straightforward to integrate because they contained consistent and
largely homogeneous variable names. We therefore retained the same
variables from each dataset, allowing us to make meaningful comparisons
in food access between 2015 and 2019.

# Ethics, Privacy & Security

**Ethics**

This project involves data describing vulnerable populations,
specifically seniors aged 65 and over and SNAP-receiving households.
These populations require discretion as careless findings or framing
about them can further stigmatise affected communities. A key
consideration for this analysis is avoiding deficit oriented framing,
practically, this means describing a tract as “low access” reflects a
structural and geographic outcome specifically the distance for a
household to supermarkets or transport infrastructure. It is not
intended to signify the choices of the individuals living in the
household, this is to avoid perpetuating any stereotypes or stigmas. As
a result the language of the report has been mindfully selected to
reflect this. For example we would state “A tract has a high rate of
low-access seniors” instead of language that could be read as describing
a senior or snap receiving households like “seniors in this area don’t
shop at super markets”

Another ethical risk we are considering is an ecological fallacy. All
analysis in this project is conducted at the census-tract level which
uses aggregated rates for an area rather than individual data. So where
a tract is classified as “low access” indicates the aggregate population
meets USDA distance thresholds, it does not mean that every individual
within that tract lacks access. As a result our conclusions are based on
patterns observed in tracts, not on individuals.

**Privacy**

This project uses only publicly available, government-released data
(USDA Food Access Research Atlas and U.S. Census Bureau data), both
provided at the census-tract level. We did no original data collection
of our own, or have any interaction with the individuals involved with
the data. The data is already aggregated tract-areas and released for
public use as verified by the U.S. government so the privacy risk to
individuals is minimal. The risks too to privacy are not technical with
how the data is structured or analysed but in the interpretation of the
data. There is still a risk (although very small risk) of individual
identification for tracts with very small populations e.g a small amount
of senior SNAP households in a rural tract, where in principle it could
be easier to infer information about specific households. This
represents a limitation of the dataset from a privacy perspective,
though it is a low and indirect risk rather than a risk introduced by
our analysis.

**Security**

As mentioned in the privacy section this project uses only public,
pre-aggregated data. The overall security risk associated with this
project is low. We practiced secure data management in this project
where we store all our working data and code within a shared private
Github repository, accessible only to the members of our group as
opposed to a public repository or sharing code in any other external
source. This gave us a controlled and access-limited location for our
data, code and analysis. With some additional time and resources to
allocate into security it would be worthwhile establishing a documented
data deletion plan for after the project concludes.

```{r 2015 load, echo = FALSE}
library(readxl)
DATA2015 <- read_excel("FoodAccessResearchAtlasData2015.xlsx", 
                       sheet = 3, na = c("", "NULL"))
```

```{r 2019 load, echo = FALSE}
DATA2019 <- read_excel("FoodAccessResearchAtlasData2019.xlsx", 
                       sheet = 3, na = c("", "NULL"))
```

```{r, echo = FALSE, include=FALSE}
dim(DATA2015)
dim(DATA2019)
names(DATA2015)
names(DATA2019)
str(DATA2015)
str(DATA2019)
```

```{r, echo = FALSE, include=FALSE}
# To Unify variable names
names(DATA2015) <- tolower(names(DATA2015))
names(DATA2019) <- tolower(names(DATA2019))
```

```{r, include=FALSE}
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
```

```{r, echo = FALSE}
# Select variables we need
data2015_clean <- DATA2015 %>%
  select(
    censustract, state, county, urban, pop2010, ohu2010, laseniors1, laseniors10, lasnap1, laseniors1share, laseniors10share, lasnap1share, lasnap10share, lasnap10, laseniors20, laseniorshalf, laseniorshalfshare, tractseniors, tractsnap
  )

data2019_clean <- DATA2019 %>%
  select(
    censustract, state, county, urban, pop2010, ohu2010, laseniors1, laseniors10, laseniors1share, laseniors10share, lasnap1share, lasnap10share, lasnap1, lasnap10, laseniors20, laseniorshalf, laseniorshalfshare, tractseniors, tractsnap, 
)
```

```{r, echo = FALSE, include=FALSE}
numeric_vars <- c(
  "urban", "pop2010", "ohu2010", "laseniors1", "laseniors10", "laseniors1share", "laseniors10share", "lasnap1", "lasnap10", "lasnap1share", "lasnap10share", "tractseniors", "tractsnap"
)

data2015_clean <- data2015_clean %>%
  mutate(
    across(
      any_of(numeric_vars),
      as.numeric
    )
  )

data2019_clean <- data2019_clean %>%
  mutate(
    across(
      any_of(numeric_vars),
      as.numeric
    )
  )

combined_data <- bind_rows(
  data2015_clean,
  data2019_clean
)

combined_data
```

```{r, echo = FALSE, include=FALSE}
sapply(
  data2015_clean[numeric_vars],
  class
)

sapply(
  data2019_clean[numeric_vars],
  class
)
```

```{r, echo = FALSE}
data2015_clean <- data2015_clean %>%
  mutate(year = 2015)

data2019_clean <- data2019_clean %>%
  mutate(year = 2019)
```

```{r Urban / rural Classifier, echo = FALSE}
#Create Urban/Rural
data2015_clean <- data2015_clean %>%
  mutate(
    year = 2015,
    area = if_else(
      urban == 1,
      "Urban",
      "Rural",
      missing = NA_character_
    )
  )

data2019_clean <- data2019_clean %>%
  mutate(
    year = 2019,
    area = if_else(
      urban == 1,
      "Urban",
      "Rural",
      missing = NA_character_
    )
  )
```

```{r, echo = FALSE, include=FALSE}
# 2015 of senior and snap
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_n = case_when(
      urban == 1 ~ laseniors1,
      urban == 0 ~ laseniors10,
      TRUE ~ NA_real_
    ),

    snap_la_n = case_when(
      urban == 1 ~ lasnap1,
      urban == 0 ~ lasnap10,
      TRUE ~ NA_real_
    )
  )

# 2019 of senior and snap
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_n = case_when(
      urban == 1 ~ laseniors1,
      urban == 0 ~ laseniors10,
      TRUE ~ NA_real_
    ),

    snap_la_n = case_when(
      urban == 1 ~ lasnap1,
      urban == 0 ~ lasnap10,
      TRUE ~ NA_real_
    )
  )
```

```{r, echo = FALSE, include=FALSE}
# Show precentage
# 2015
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_pct = case_when(
      urban == 1 ~ laseniors1share * 100,
      urban == 0 ~ laseniors10share * 100,
      TRUE ~ NA_real_
    ),

    snap_la_pct = case_when(
      urban == 1 ~ lasnap1share * 100,
      urban == 0 ~ lasnap10share * 100,
      TRUE ~ NA_real_
    )
  )

# 2019
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_pct = case_when(
      urban == 1 ~ laseniors1share,
      urban == 0 ~ laseniors10share,
      TRUE ~ NA_real_
    ),

    snap_la_pct = case_when(
      urban == 1 ~ lasnap1share,
      urban == 0 ~ lasnap10share,
      TRUE ~ NA_real_
    )
  )
```

```{r, echo = FALSE, include=FALSE}
# Seniors
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_n = if_else(
      urban == 1,
      laseniors1,
      laseniors10
    )
  )

# SNAP
data2015_clean <- data2015_clean %>%
  mutate(
    snap_la_n = if_else(
      urban == 1,
      lasnap1,
      lasnap10
    )
  )

# Seniors
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_n = if_else(
      urban == 1,
      laseniors1,
      laseniors10
    )
  )

# SNAP
data2019_clean <- data2019_clean %>%
  mutate(
    snap_la_n = if_else(
      urban == 1,
      lasnap1,
      lasnap10
    )
  )
```

```{r, echo = FALSE, include=FALSE}
summary(data2015_clean)
summary(data2019_clean)
```

```{r, echo = FALSE, include=FALSE}
#data2019_clean <- data2019_clean %>%
#  mutate(
#    across(
#      c(laseniors1, laseniors10, lasnap1, lasnap10,
#        laseniors20,laseniorshalf,laseniorshalfshare,senior_la_n, snap_la_n),
#      ~ as.numeric(ifelse(.x == "NULL", "0", .x))
#    )
#  )
```

```{r, echo = FALSE, include=FALSE}
summary(data2019_clean)
```

```{r, echo = FALSE, include=FALSE}
library(knitr)

dataset_table <- data.frame(
  Dataset = c("2015 Food Access Research Atlas",
              "2019 Food Access Research Atlas"),
  Observations = c(72864, 72531),
  Variables = c(147, 147),
  Unit_of_Analysis = c("Census tract", "Census tract"),
  Main_Data_Types = c(
    "Geographic identifiers, urban/rural indicator, population counts, food-access measures",
    "Geographic identifiers, urban/rural indicator, population counts, food-access measures"
  )
)

kable(
  dataset_table,
  caption = "Structure of the 2015 and 2019 Food Access Research Atlas datasets",
  col.names = c(
    "Dataset",
    "Observations",
    "Variables",
    "Unit of Analysis",
    "Main Data Types"
  ),
  align = c("l", "r", "r", "l", "l")
)
```

```{r, echo = FALSE, include=FALSE}
summary_stats <- bind_rows(
  
  data2015_clean %>%
    transmute(
      year = 2015,
      Senior = senior_la_pct,
      SNAP = snap_la_pct
    ),
  
  data2019_clean %>%
    transmute(
      year = 2019,
      Senior = senior_la_pct,
      SNAP = snap_la_pct
    )
  
) %>%
  
  pivot_longer(
    cols = c(Senior, SNAP),
    names_to = "Group",
    values_to = "Low_Access_Pct"
  ) %>%
  
  group_by(year, Group) %>%
  
  summarise(
    Min = min(Low_Access_Pct, na.rm = TRUE),
    Q1 = quantile(Low_Access_Pct, 0.25, na.rm = TRUE),
    Median = median(Low_Access_Pct, na.rm = TRUE),
    Mean = mean(Low_Access_Pct, na.rm = TRUE),
    Q3 = quantile(Low_Access_Pct, 0.75, na.rm = TRUE),
    Max = max(Low_Access_Pct, na.rm = TRUE),
    Missing = sum(is.na(Low_Access_Pct)),
    .groups = "drop"
  ) %>%
  
  mutate(
    across(
      c(Min, Q1, Median, Mean, Q3, Max),
      ~ round(.x, 2)
    )
  )

summary_stats
```

```{r, echo = FALSE, include=FALSE}
kable(
  summary_stats,
  caption = "Summary Statistics for Senior and SNAP Low-Access Shares, 2015 and 2019",
  col.names = c(
    "Year", "Group", "Min (%)", "Q1 (%)", "Median (%)", "Mean (%)", "Q3 (%)", "Max (%)", "Missing"
  ),
  align = c(
    "c", "l",
    "r", "r", "r", "r", "r", "r", "r"
  )
)
```

```{r, echo = FALSE, include=FALSE}
# Check whether over 100
data2015_clean %>%
  summarise(
    senior_over_100 = sum(senior_la_pct > 100, na.rm = TRUE),
    snap_over_100 = sum(snap_la_pct > 100, na.rm = TRUE)
  )

data2019_clean %>%
  summarise(
    senior_over_100 = sum(senior_la_pct > 100, na.rm = TRUE),
    snap_over_100 = sum(snap_la_pct > 100, na.rm = TRUE)
  )
```

```{r, echo = FALSE, include=FALSE}
data2015_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )

data2019_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )
```

```{r, echo = FALSE, include=FALSE}
options(digits = 15)

data2015_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )

data2019_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )
```

```{r, echo = FALSE, include=FALSE}
combined_data <- bind_rows(
  data2015_clean,
  data2019_clean
)
combined_data
```

```{r, echo = FALSE, include=FALSE}
dim(combined_data)
table(combined_data$year)
table(combined_data$area, useNA = "ifany")
```

After harmonisation, the 2015 and 2019 datasets were combined into a
dataset containing 145,395 census-tract observations. Of these, 110,341
were classified as urban and 35,054 as rural, with no missing values in
the urban–rural classification.

```{r, echo = FALSE, include=FALSE}
summary_table <- combined_data %>%
  group_by(year, area) %>%
  summarise(
    n_tracts = n(),

    n_valid_senior =
      sum(!is.na(senior_la_pct)),

    n_valid_snap =
      sum(!is.na(snap_la_pct)),

    mean_senior_pct =
      mean(senior_la_pct, na.rm = TRUE),

    median_senior_pct =
      median(senior_la_pct, na.rm = TRUE),

    mean_snap_pct =
      mean(snap_la_pct, na.rm = TRUE),

    median_snap_pct =
      median(snap_la_pct, na.rm = TRUE),

    .groups = "drop"
  )

summary_table
```

```{r, echo = FALSE, include=FALSE}
summary_table_report <- summary_table %>%
  mutate(
    across(
      c(
        mean_senior_pct,
        median_senior_pct,
        mean_snap_pct,
        median_snap_pct
      ),
      ~ round(.x, 2)
    )
  )

summary_table_report
```

```{r, echo = FALSE, include=FALSE}
dim(data2015_clean)
names(data2015_clean)
str(data2015_clean)
summary(data2015_clean)
```

```{r, echo = FALSE, include=FALSE}
dim(data2019_clean)
names(data2019_clean)
str(data2019_clean)
summary(data2019_clean)
```

```{r, echo = FALSE, include=FALSE}
colSums(is.na(data2019_clean))
```

```{r, echo = FALSE, include=FALSE}
colSums(is.na(data2015_clean))
```

# Exploratory Data Analysis

## 3.1 Overview of the Analysis

The exploratory analysis focused on the low-supermarket-access shares
for adults aged 65 years and over and SNAP-receiving households.
Comparisons were made between 2015 and 2019 and between urban and rural
census tracts. For urban census tracts, low supermarket access was
measured using the one-mile threshold, while the ten-mile threshold was
used for rural census tracts. All low-access shares were expressed as
percentages to ensure consistency between the 2015 and 2019 datasets.

Table 1 summarises the mean and median low-access shares by year and
urban/rural status. Urban census tracts generally had higher low-access
shares than rural census tracts for both seniors and SNAP households. In
addition, the mean was consistently higher than the median, suggesting
that the distributions were right-skewed. More detailed distributions
are examined in the following sections.

```{r}
summary_table <- combined_data %>%
  group_by(area, year) %>%
  summarise(
    Senior_Mean   = round(mean(senior_la_pct, na.rm = TRUE), 1),
    Senior_Median = round(median(senior_la_pct, na.rm = TRUE), 1),
    SNAP_Mean     = round(mean(snap_la_pct, na.rm = TRUE), 1),
    SNAP_Median   = round(median(snap_la_pct, na.rm = TRUE), 1),
    n_tracts      = n(),
    .groups = "drop"
  )

kable(summary_table,
      caption = "Mean and median low-access share (%) for seniors and SNAP households, by area and year.",
      col.names = c("Area", "Year", "Senior Mean", "Senior Median",
                    "SNAP Mean", "SNAP Median", "N Tracts"))
```

Table 2 shows substantial differences in data completeness between the
two years. There were no missing values in the Senior or SNAP low-access
share variables used for 2015, whereas the 2019 data contained 29,953
missing senior observations and 29,918 missing SNAP observations.
Therefore, comparisons involving 2019 should be interpreted with caution
because they are based only on census tracts with available low-access
information.

```{r}
missing_table <- data.frame(
  Year = c(2015, 2019),
  Senior_Missing = c(
    sum(is.na(data2015_clean$senior_la_pct)),
    sum(is.na(data2019_clean$senior_la_pct))
  ),
  SNAP_Missing = c(
    sum(is.na(data2015_clean$snap_la_pct)),
    sum(is.na(data2019_clean$snap_la_pct))
  )
)

kable(missing_table,
      caption = "Number of missing Senior and SNAP low-access share observations, by year.",
      col.names = c("Year", "Senior Missing", "SNAP Missing"))
```

## 3.2 Senior Low-Access Share: Distribution and Urban–Rural Comparison

Figure 1 shows the distribution of senior low-access shares by year and
urban/rural status. The distributions were strongly right-skewed across
all four groups, with most census tracts having relatively low senior
low-access shares and a smaller number having much higher values. Urban
census tracts generally showed higher low-access shares than rural
census tracts. The 2019 distributions also appeared shifted toward
higher values than in 2015. However, interpret the 2019 results
cautiously because the senior low-access variable contained substantial
missing data.

```{r}
ggplot(combined_data, aes(x = senior_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  facet_grid(area ~ year) +
  labs(
    title = "Figure 1: Distribution of Senior Low-Access Share",
    x = "Senior Low-Access Share (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

Figure 2 provides a more direct comparison between urban and rural
census tracts. Urban tracts had higher median senior low-access shares
than rural tracts in both 2015 and 2019, and the 2019 distributions were
generally higher than those in 2015. Both years also contained several
high-value outliers, which is consistent with the right-skewed
distributions observed in Figure 1.

```{r}
# Senior Boxplot
ggplot(
  combined_data,
  aes(
    x = factor(year),
    y = senior_la_pct,
    fill = area
  )
) +
  geom_boxplot(
    outlier.size = 0.3,
    na.rm = TRUE
  ) +
  scale_fill_grey() +
  labs(
    title = "Figure 2: Senior Low-Access Share by Year and Area",
    x = "Year",
    y = "Senior Low-Access Share (%)",
    fill = "Area"
  ) +
  theme_minimal()
```

Overall, the senior low-access share was generally higher in urban than
rural census tracts, and the distributions were strongly right-skewed in
both years. The apparent differences between 2015 and 2019 should be
interpreted cautiously because of substantial missing data in 2019.

## 3.3 SNAP Low-Access Share: Distribution and Urban–Rural Comparison

Figure 3 shows the distribution of SNAP low-access shares across census
tracts. As with the senior measure, the distribution is right-skewed,
with many census tracts concentrated at relatively low percentages and a
smaller number of tracts having much higher values.

```{r}
ggplot(combined_data, aes(x = snap_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(
    title = "Figure 3: Distribution of SNAP Low-Access Share, All Years and Areas",
    x = "SNAP Low-Access Share (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

Figure 4 compares the distribution of SNAP low-access shares between
urban and rural census tracts in 2015 and 2019. The plot allows
differences in median, spread and extreme values to be compared while
also showing whether the urban–rural pattern remained similar across the
two years.

```{r}
ggplot(combined_data,aes(x = snap_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  facet_grid(area ~ year) +
  labs(
    title = "Figure 4: Distribution of SNAP Low-Access Share",
    x = "SNAP Low-Access Share (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

Overall, SNAP low-access shares were generally higher in urban than
rural census tracts, with strongly right-skewed distributions in both
years. The apparent differences between 2015 and 2019 should be
interpreted cautiously because of substantial missing data in 2019.

## 3.4 Changes between 2015 and 2019

To examine changes between 2015 and 2019 more directly, census tracts in
the two datasets were matched using their Census Tract identifiers.

```{r, echo = FALSE, include = FALSE}
# Match the same census tract in 2015 and 2019, and calculate the changes in the share of seniors and SNAP low-access individuals between the two years.

matched_data <- data2015_clean %>%
  select(
    censustract,
    area_2015 = area,
    senior_pct_2015 = senior_la_pct,
    snap_pct_2015 = snap_la_pct
  ) %>%
  inner_join(
    data2019_clean %>%
      select(
        censustract,
        area_2019 = area,
        senior_pct_2019 = senior_la_pct,
        snap_pct_2019 = snap_la_pct
      ),
    by = "censustract"
  ) %>%
  mutate(
    senior_change = senior_pct_2019 - senior_pct_2015,
    snap_change = snap_pct_2019 - snap_pct_2015
  )

matched_data
```

```{r, echo = FALSE, include = FALSE}
dim(matched_data)
sum(!is.na(matched_data$senior_change))
sum(!is.na(matched_data$snap_change))
```

A total of 72,506 census tracts were successfully matched across the two
years. Of these, 21 census tracts changed their urban/rural
classification between 2015 and 2019. These tracts were excluded from
the change analysis because the low-access distance threshold differs
between urban and rural areas. This resulted in 72,485 census tracts
with a stable urban/rural classification.

```{r}
# Senior Change (using all matched tracts, before excluding reclassified tracts)
ggplot(
  matched_data,
  aes(x = area_2015, y = senior_change)) +
  geom_boxplot(na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Change in Senior Low-Access Share, 2015-2019",
    x = "Area",
    y = "Change in percentage points"
  ) +
  theme_minimal()
```

```{r, echo = FALSE, include = FALSE}
# Check the boxplot why is the line 
matched_data %>%
  group_by(area_2015) %>%
  summarise(n = n(), 
            Q1 = quantile(senior_change, 0.25, na.rm = TRUE),
            Median = median(senior_change, na.rm = TRUE),
            Q3 = quantile(senior_change, 0.75, na.rm = TRUE),
            Min = min(senior_change, na.rm = TRUE),
            Max = max(senior_change, na.rm = TRUE),
            pct_exact_zero =mean(senior_change == 0,na.rm = TRUE) * 100
  )

```

```{r, echo = FALSE, include = FALSE}
matched_data %>%
  group_by(area_2015) %>%
  summarise(
    pct_near_zero =
      mean(
        abs(senior_change) < 1e-8,
        na.rm = TRUE
      ) * 100
  )
```

Reason for why choose histogram to show the Change in Senior Low-Access
Share between 2015 and 2019: Based on the summary, we find the boxplot
is right. However, the vast majority of changes in census tracts are
very close to zero. From the summary, especially Urban, Q1 = -4.019 x
10\^-14 Median = 0 Q3 = 4.974 x 10\^-14 The real meaning of these three
number is 0% points. Thus, the boxplot of Urban: Q1 $\approx$ Median
$\approx$ Q3 $\approx$ 0, it shows the box was naturally compressed into
a line near y=0.

While, Rural is not completely 0: Q1 $\approx$ -0.014 Median = 0 Q3
$\approx$ 0.052 However, the IQR just around 0.052-(-0.014) $\approx$
0.066. There is the very small percentage points. Compare the
Min(-38.16) and Max(+40.72), so the vertical axis is stretched very wide
to accommodate these extreme changes, making the box even less visible.

The key finding We cannot be fully trust "pct_exact_zero". You now
obtain: Rural: 10.41% Urban: 15.50% It seems that only 10–15% have
remained unchanged.

However, for Urban, Q1 and Q3 are: -0.000000000000040 +0.000000000000050

This is actually floating-point numerical noise. R will consider:
0.00000000000005 != 0 So you use: senior_change == 0

It will classify many observations that are actually equivalent to 0 as
"non-zero changes".

```{r}
# SNAP
ggplot(
  matched_data,
  aes(x = area_2015, y = snap_change)) +
  geom_boxplot(na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Change in SNAP Low-Access Share, 2015-2019",
    x = "Area",
    y = "Change in percentage points"
  ) +
  theme_minimal()
```

```{r, echo = FALSE, include = FALSE}
matched_stable <- matched_data %>%
  filter(area_2015 == area_2019)

matched_stable
```

```{r}
matched_stable %>%
  group_by(area_2015) %>%
  summarise(
    n = sum(!is.na(snap_change)),
    Q1 = quantile(snap_change, 0.25, na.rm = TRUE),
    Median = median(snap_change, na.rm = TRUE),
    Q3 = quantile(snap_change, 0.75, na.rm = TRUE),
    Min = min(snap_change, na.rm = TRUE),
    Max = max(snap_change, na.rm = TRUE),
    .groups = "drop"
  )

matched_stable %>%
  group_by(area_2019) %>%
  summarise(
    n = sum(!is.na(snap_change)),
    Q1 = quantile(snap_change, 0.25, na.rm = TRUE),
    Median = median(snap_change, na.rm = TRUE),
    Q3 = quantile(snap_change, 0.75, na.rm = TRUE),
    Min = min(snap_change, na.rm = TRUE),
    Max = max(snap_change, na.rm = TRUE),
    .groups = "drop"
  )
```

Still check, Q1, Median, and Q3 all close to 0, it shows the data itself
is highly concentrated around 0.

```{r}
ggplot(matched_data,aes(x = senior_change)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  facet_wrap(~ area_2015) +
  geom_vline(xintercept = 0,linetype = "dashed") +
  labs(
    title = "Distribution of Changes in Senior Low-Access Share, 2015-2019 (unfiltered)",
    x = "Change in percentage points",
    y = "Density"
  ) +
  theme_minimal()
```

This will show more clearly than boxplot: A large number of tracts are
concentrated around 0; Which ones are increases? Which ones are
reductions? Is the distribution of Urban and Rural areas different?

```{r, echo = FALSE, include = FALSE}
change_summary <- matched_data %>%
  group_by(area_2015) %>%
  summarise(
    n = sum(!is.na(senior_change)),
    median_change =median(senior_change, na.rm = TRUE),
    mean_change =mean(senior_change, na.rm = TRUE),
    pct_increase =mean(senior_change > 0.1, na.rm = TRUE) * 100,
    pct_decrease =mean(senior_change < -0.1, na.rm = TRUE) * 100,
    pct_stable = mean(abs(senior_change) <= 0.1, na.rm = TRUE) * 100,
    .groups = "drop"
  )

change_summary
```

```{r, echo = FALSE, include = FALSE}
table(
  matched_data$area_2015,
  matched_data$area_2019
)
```

We should use the second which better than last one. Even though we
cannot see observable difference, it shows the results which checked:
Rural to Urban: 10; Urban to Rural: 11 . There are 21 census tracts
deleted.

Therefore, although the second graph does not show any visual
difference, it is more rigorous in its approach because it ensures that:
For rural areas, the comparison always uses a 10-mile measure, and for
urban areas, the comparison always uses a 1-mile measure. So, the
subsequent 2015 - 2019 change analysis suggests using "matched_stable".

```{r, echo = FALSE, include = FALSE}
# Make sure the number of row between two datasets
nrow(matched_data)
nrow(matched_stable)

nrow(matched_data) - nrow(matched_stable)
```

```{r, echo = FALSE, include = FALSE}
matched_stable <- matched_stable %>%
  mutate(snap_change = snap_pct_2019 - snap_pct_2015)

matched_stable
```

For the paired analysis, only census tracts with non-missing low-access
shares in both 2015 and 2019 were included. Change was calculated as:
Change= Low-access share2019 - Low-access share2015 and is reported in
percentage points. Positive values therefore represent an increase in
low-access share between 2015 and 2019, while negative values represent
a decrease.

```{r, echo = FALSE, include = FALSE}
senior_paired_data <- matched_stable %>%
  filter(
    !is.na(senior_pct_2015),
    !is.na(senior_pct_2019)
  )

senior_paired <- senior_paired_data %>%
  group_by(area_2015) %>%
  summarise(
    N = n(),
    
    Mean_2015 = mean(senior_pct_2015),
    Mean_2019 = mean(senior_pct_2019),
    
    Median_2015 = median(senior_pct_2015),
    Median_2019 = median(senior_pct_2019),
    
    Mean_Change = mean(senior_change),
    Median_Change = median(senior_change),
    
    .groups = "drop"
  ) %>%
  mutate(
    across(
      c(
        Mean_2015,
        Mean_2019,
        Median_2015,
        Median_2019,
        Mean_Change,
        Median_Change
      ),
      ~ round(.x, 2)
    )
  )

kable(
  senior_paired,
  caption = "Paired summary of Senior low-access share, 2015-2019",
  col.names = c(
    "Area",
    "N",
    "Mean 2015 (%)",
    "Mean 2019 (%)",
    "Median 2015 (%)",
    "Median 2019 (%)",
    "Mean Change (pp)",
    "Median Change (pp)"
  )
)
```

```{r, echo = FALSE, include = FALSE}
snap_paired_data <- matched_stable %>%
  filter(
    !is.na(snap_pct_2015),
    !is.na(snap_pct_2019)
  )

snap_paired <- snap_paired_data %>%
  group_by(area_2015) %>%
  summarise(
    N = n(),
    
    Mean_2015 = mean(snap_pct_2015),
    Mean_2019 = mean(snap_pct_2019),
    
    Median_2015 = median(snap_pct_2015),
    Median_2019 = median(snap_pct_2019),
    
    Mean_Change = mean(snap_change),
    Median_Change = median(snap_change),
    
    .groups = "drop"
  ) %>%
  mutate(
    across(
      c(
        Mean_2015,
        Mean_2019,
        Median_2015,
        Median_2019,
        Mean_Change,
        Median_Change
      ),
      ~ round(.x, 2)
    )
  )

kable(
  snap_paired,
  caption = "Paired summary of SNAP low-access share, 2015-2019",
  col.names = c(
    "Area",
    "N",
    "Mean 2015 (%)",
    "Mean 2019 (%)",
    "Median 2015 (%)",
    "Median 2019 (%)",
    "Mean Change (pp)",
    "Median Change (pp)"
  )
)
```

```{r}
paired_summary <- bind_rows(
  senior_paired %>%
    mutate(Group = "Senior"),
  
  snap_paired %>%
    mutate(Group = "SNAP")
) %>%
  select(
    Group,
    area_2015,
    N,
    Mean_2015,
    Mean_2019,
    Median_2015,
    Median_2019,
    Mean_Change,
    Median_Change
  ) %>%
  arrange(Group, area_2015)

kable(
  paired_summary,
  caption = "Paired summary of Senior and SNAP low-access shares, 2015-2019",
  col.names = c(
    "Group",
    "Area",
    "N",
    "Mean 2015 (%)",
    "Mean 2019 (%)",
    "Median 2015 (%)",
    "Median 2019 (%)",
    "Mean Change (pp)",
    "Median Change (pp)"
  ),
  align = c(
    "l", "l", "r",
    "r", "r", "r", "r", "r", "r"
  )
)
```

From Table 3, we compared the different groups with the different areas
between 2015 and 2019. For seniors, the paired mean low-access share
increased only slightly between 2015 and 2019. In rural census tracts,
the mean increased from 3.94% to 3.96%, corresponding to an average
increase of 0.02 percentage points. In urban census tracts, the mean
increased from 5.21% to 5.32%, an average increase of 0.11 percentage
points. The median change was zero in both rural and urban tracts,
indicating that the typical matched census tract experienced little or
no change. For SNAP households, the direction of the average change
differed from that observed for seniors. The mean low-access share
decreased from 2.91% to 2.77% in rural census tracts, corresponding to a
mean change of -0.14 percentage points. In urban census tracts, the mean
decreased from 4.40% to 4.33%, corresponding to a mean change of -0.07
percentage points. However, the median change was again zero in both
groups, suggesting that changes for a typical matched census tract were
very small.



Figure 5 shows that changes in senior low-access share were strongly
concentrated around zero in both rural and urban census tracts. This is
consistent with the zero median changes reported in Table 3. Although
most census tracts experienced relatively small changes, a smaller
number showed substantially larger positive or negative changes.

```{r}
ggplot(matched_stable, aes(x = senior_change)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  facet_wrap(~ area_2015) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Figure 5: Distribution of Changes in Senior Low-Access Share, 2015-2019",
    x = "Change in percentage points",
    y = "Density"
  ) +
  theme_minimal()
```

Figure 6 shows that change distributions in SNAP Low-Access share were
also strongly concentrated around zero, although some census tracts
experienced larger increases or decreases. The concentration around zero
is consistent with the paired summary statistics and indicates that the
overall changes between the two years were small for most matched census
tracts.

```{r}
ggplot(matched_stable,aes(x = snap_change)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  facet_wrap(~ area_2015) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Figure 6: Distribution of Changes in SNAP Low-Access Share, 2015-2019",
    x = "Change in percentage points",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

```{r}
matched_stable %>%
  group_by(area_2015) %>%
  summarise(
    n = sum(!is.na(snap_change)),
    Q1 = quantile(snap_change,0.25,na.rm = TRUE),
    Median = median(snap_change,na.rm = TRUE),
    Q3 = quantile(snap_change,0.75,na.rm = TRUE),
    Min = min(snap_change,na.rm = TRUE),
    Max = max(snap_change,na.rm = TRUE),
    pct_near_zero =mean(abs(snap_change) < 1e-8,na.rm = TRUE) * 100,
    .groups = "drop"
  )
```

Overall, the paired analysis suggests that tract-level changes between
2015 and 2019 were generally small. Senior low-access shares showed
slight positive average changes in both rural and urban areas, whereas
SNAP low-access shares showed slight negative average changes. However,
median changes were zero for all four comparisons, indicating that the
mean changes were influenced by a smaller number of census tracts with
larger positive or negative changes. These findings are descriptive and
do not by themselves establish whether the observed differences are
statistically significant.

## 3.5 Additional Comparison: Mean Low-Access Share by Group

To talk about... median much lower then means = skewed data. Urban
vastly outnumber rural. Values increase across all four rows

keeping in mind a rural tract requires households and seniors to be at
least 10 miles away compared to a min of 1

```{r}
library(tidyr)

dot_data <- summary_table %>%
  select(area, year, Senior_Mean, SNAP_Mean) %>%
  pivot_longer(cols = c(Senior_Mean, SNAP_Mean),
               names_to = "Group", values_to = "Mean_Rate") %>%
  mutate(Group = recode(Group, Senior_Mean = "Seniors", SNAP_Mean = "SNAP"))

ggplot(dot_data, aes(x = Mean_Rate, y = Group, color = factor(year))) +
  geom_point(size = 3) +
  facet_wrap(~area) +
  labs(
    title = "Mean Low-Access Share by Group, Area, and Year",
    x = "Mean Low-Access Share (%)",
    y = NULL,
    color = "Year"
  ) +
  theme_minimal()
```

Across both Senior and SNAP low-access shares, 2019 shows a higher mean
rate than 2015, and Urban tracts show a higher mean rate than Rural
tracts, in both years and for both groups

Because "low-access" is defined using a much stricter distance threshold
in urban areas (1 mile) than rural areas (10 miles), a higher urban
low-access rate does not necessarily mean urban residents travel further
to reach a supermarket than rural residents. This may just reflect that
urban areas are held to a stricter access standard. A rural resident
traveling several miles to a store may still count as "having access"
under the rural threshold, while a much shorter distance in a city would
count as "low access."

As a result, the lower mean low-access share in rural areas should not
be read as evidence that rural food access is genuinely better. Rural
residents may need to travel considerably farther in absolute terms
before being classified as low-access at all, meaning the rural rate
likely understates the true burden of distance compared to the urban
rate. This distinction should be kept in mind whenever urban and rural
low-access rates are compared directly in this report, rather than
treating them as measuring the same underlying concept.

## 3.6 Summary

The exploratory analysis identified several important patterns in the
senior and SNAP low-access measures. First, both low-access share
distributions were strongly right-skewed. Most census tracts had
relatively low low-access shares, while a smaller number of tracts had
substantially higher values. As a result, the mean was generally higher
than the median, indicating that the mean was influenced by census
tracts with relatively large low-access shares.

Although the 2019 cross-sectional distributions appeared higher than
those in 2015, these comparisons should be interpreted cautiously
because the 2019 data contained substantial missing values. The paired
analysis provided a more direct comparison of the same census tracts
across years and showed that changes were generally small. Senior
low-access shares increased slightly on average, while SNAP low-access
shares decreased slightly, and the median change was zero across both
rural and urban tracts.
