---
title: "FoodAccessPREP"
author: "Oliver Temel, Mairangi Campbell and Jiayi Lu"
IDs: " , , 300633898" 
date: "2026-09-03"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r 2015 load, echo = FALSE, include = FALSE}
library(readxl)
DATA2015 <- read_excel("FoodAccessResearchAtlasData2015.xlsx", 
                       sheet = 3, na = c("", "NULL"))
```

```{r 2019 load, echo = FALSE, include = FALSE}
DATA2019 <- read_excel("FoodAccessResearchAtlasData2019.xlsx", 
                       sheet = 3, na = c("", "NULL"))
```

```{r, echo = FALSE, include=FALSE}
dim(DATA2015)
dim(DATA2019)
names(DATA2015)
names(DATA2019)
str(DATA2015)
str(DATA2019)
```

```{r, echo = FALSE, include=FALSE}
# To Unify variable names
names(DATA2015) <- tolower(names(DATA2015))
names(DATA2019) <- tolower(names(DATA2019))
```

```{r, include=FALSE}
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
```

```{r, echo = FALSE, include = FALSE}
# Select variables we need
data2015_clean <- DATA2015 %>%
  select(
    censustract, state, county, urban, pop2010, ohu2010, laseniors1, laseniors10, lasnap1, laseniors1share, laseniors10share, lasnap1share, lasnap10share, lasnap10, laseniors20, laseniorshalf, laseniorshalfshare, tractseniors, tractsnap
  )

data2019_clean <- DATA2019 %>%
  select(
    censustract, state, county, urban, pop2010, ohu2010, laseniors1, laseniors10, laseniors1share, laseniors10share, lasnap1share, lasnap10share, lasnap1, lasnap10, laseniors20, laseniorshalf, laseniorshalfshare, tractseniors, tractsnap, 
)
```

```{r, echo = FALSE, include=FALSE}
numeric_vars <- c(
  "urban", "pop2010", "ohu2010", "laseniors1", "laseniors10", "laseniors1share", "laseniors10share", "lasnap1", "lasnap10", "lasnap1share", "lasnap10share", "tractseniors", "tractsnap"
)

data2015_clean <- data2015_clean %>%
  mutate(
    across(
      any_of(numeric_vars),
      as.numeric
    )
  )

data2019_clean <- data2019_clean %>%
  mutate(
    across(
      any_of(numeric_vars),
      as.numeric
    )
  )

combined_data <- bind_rows(
  data2015_clean,
  data2019_clean
)

combined_data
```

```{r, echo = FALSE, include=FALSE}
sapply(
  data2015_clean[numeric_vars],
  class
)

sapply(
  data2019_clean[numeric_vars],
  class
)
```

```{r, echo = FALSE, include = FALSE}
data2015_clean <- data2015_clean %>%
  mutate(year = 2015)

data2019_clean <- data2019_clean %>%
  mutate(year = 2019)
```

```{r Urban / rural Classifier, echo = FALSE}
#Create Urban/Rural
data2015_clean <- data2015_clean %>%
  mutate(
    year = 2015,
    area = if_else(
      urban == 1,
      "Urban",
      "Rural",
      missing = NA_character_
    )
  )

data2019_clean <- data2019_clean %>%
  mutate(
    year = 2019,
    area = if_else(
      urban == 1,
      "Urban",
      "Rural",
      missing = NA_character_
    )
  )
```

```{r, echo = FALSE, include=FALSE}
# 2015 of senior and snap
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_n = case_when(
      urban == 1 ~ laseniors1,
      urban == 0 ~ laseniors10,
      TRUE ~ NA_real_
    ),

    snap_la_n = case_when(
      urban == 1 ~ lasnap1,
      urban == 0 ~ lasnap10,
      TRUE ~ NA_real_
    )
  )

# 2019 of senior and snap
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_n = case_when(
      urban == 1 ~ laseniors1,
      urban == 0 ~ laseniors10,
      TRUE ~ NA_real_
    ),

    snap_la_n = case_when(
      urban == 1 ~ lasnap1,
      urban == 0 ~ lasnap10,
      TRUE ~ NA_real_
    )
  )
```

```{r, echo = FALSE, include=FALSE}
# Show precentage
# 2015
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_pct = case_when(
      urban == 1 ~ laseniors1share * 100,
      urban == 0 ~ laseniors10share * 100,
      TRUE ~ NA_real_
    ),

    snap_la_pct = case_when(
      urban == 1 ~ lasnap1share * 100,
      urban == 0 ~ lasnap10share * 100,
      TRUE ~ NA_real_
    )
  )

# 2019
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_pct = case_when(
      urban == 1 ~ laseniors1share,
      urban == 0 ~ laseniors10share,
      TRUE ~ NA_real_
    ),

    snap_la_pct = case_when(
      urban == 1 ~ lasnap1share,
      urban == 0 ~ lasnap10share,
      TRUE ~ NA_real_
    )
  )
```

```{r, echo = FALSE, include=FALSE}
# Seniors
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_n = if_else(
      urban == 1,
      laseniors1,
      laseniors10
    )
  )

# SNAP
data2015_clean <- data2015_clean %>%
  mutate(
    snap_la_n = if_else(
      urban == 1,
      lasnap1,
      lasnap10
    )
  )

# Seniors
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_n = if_else(
      urban == 1,
      laseniors1,
      laseniors10
    )
  )

# SNAP
data2019_clean <- data2019_clean %>%
  mutate(
    snap_la_n = if_else(
      urban == 1,
      lasnap1,
      lasnap10
    )
  )
```

```{r, echo = FALSE, include=FALSE}
summary(data2015_clean)
summary(data2019_clean)
```

```{r, echo = FALSE, include=FALSE}
#data2019_clean <- data2019_clean %>%
#  mutate(
#    across(
#      c(laseniors1, laseniors10, lasnap1, lasnap10,
#        laseniors20,laseniorshalf,laseniorshalfshare,senior_la_n, snap_la_n),
#      ~ as.numeric(ifelse(.x == "NULL", "0", .x))
#    )
#  )
```

```{r, echo = FALSE, include=FALSE}
summary(data2019_clean)
```

```{r, echo = FALSE, include=FALSE}
library(knitr)

dataset_table <- data.frame(
  Dataset = c("2015 Food Access Research Atlas",
              "2019 Food Access Research Atlas"),
  Observations = c(72864, 72531),
  Variables = c(147, 147),
  Unit_of_Analysis = c("Census tract", "Census tract"),
  Main_Data_Types = c(
    "Geographic identifiers, urban/rural indicator, population counts, food-access measures",
    "Geographic identifiers, urban/rural indicator, population counts, food-access measures"
  )
)

kable(
  dataset_table,
  caption = "Structure of the 2015 and 2019 Food Access Research Atlas datasets",
  col.names = c(
    "Dataset",
    "Observations",
    "Variables",
    "Unit of Analysis",
    "Main Data Types"
  ),
  align = c("l", "r", "r", "l", "l")
)
```

```{r, echo = FALSE, include=FALSE}
summary_stats <- bind_rows(
  
  data2015_clean %>%
    transmute(
      year = 2015,
      Senior = senior_la_pct,
      SNAP = snap_la_pct
    ),
  
  data2019_clean %>%
    transmute(
      year = 2019,
      Senior = senior_la_pct,
      SNAP = snap_la_pct
    )
  
) %>%
  
  pivot_longer(
    cols = c(Senior, SNAP),
    names_to = "Group",
    values_to = "Low_Access_Pct"
  ) %>%
  
  group_by(year, Group) %>%
  
  summarise(
    Min = min(Low_Access_Pct, na.rm = TRUE),
    Q1 = quantile(Low_Access_Pct, 0.25, na.rm = TRUE),
    Median = median(Low_Access_Pct, na.rm = TRUE),
    Mean = mean(Low_Access_Pct, na.rm = TRUE),
    Q3 = quantile(Low_Access_Pct, 0.75, na.rm = TRUE),
    Max = max(Low_Access_Pct, na.rm = TRUE),
    Missing = sum(is.na(Low_Access_Pct)),
    .groups = "drop"
  ) %>%
  
  mutate(
    across(
      c(Min, Q1, Median, Mean, Q3, Max),
      ~ round(.x, 2)
    )
  )

summary_stats
```

```{r, echo = FALSE, include=FALSE}
kable(
  summary_stats,
  caption = "Summary Statistics for Senior and SNAP Low-Access Shares, 2015 and 2019",
  col.names = c(
    "Year", "Group", "Min (%)", "Q1 (%)", "Median (%)", "Mean (%)", "Q3 (%)", "Max (%)", "Missing"
  ),
  align = c(
    "c", "l",
    "r", "r", "r", "r", "r", "r", "r"
  )
)
```

```{r, echo = FALSE, include=FALSE}
# Check whether over 100
data2015_clean %>%
  summarise(
    senior_over_100 = sum(senior_la_pct > 100, na.rm = TRUE),
    snap_over_100 = sum(snap_la_pct > 100, na.rm = TRUE)
  )

data2019_clean %>%
  summarise(
    senior_over_100 = sum(senior_la_pct > 100, na.rm = TRUE),
    snap_over_100 = sum(snap_la_pct > 100, na.rm = TRUE)
  )
```

```{r, echo = FALSE, include=FALSE}
data2015_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )

data2019_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )
```

```{r, echo = FALSE, include=FALSE}
options(digits = 15)

data2015_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )

data2019_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )
```

```{r, echo = FALSE, include=FALSE}
combined_data <- bind_rows(
  data2015_clean,
  data2019_clean
)
combined_data
```

```{r, echo = FALSE, include=FALSE}
dim(combined_data)
table(combined_data$year)
table(combined_data$area, useNA = "ifany")
```

```{r, echo = FALSE, include=FALSE}
summary_table <- combined_data %>%
  group_by(year, area) %>%
  summarise(
    n_tracts = n(),

    n_valid_senior =
      sum(!is.na(senior_la_pct)),

    n_valid_snap =
      sum(!is.na(snap_la_pct)),

    mean_senior_pct =
      mean(senior_la_pct, na.rm = TRUE),

    median_senior_pct =
      median(senior_la_pct, na.rm = TRUE),

    mean_snap_pct =
      mean(snap_la_pct, na.rm = TRUE),

    median_snap_pct =
      median(snap_la_pct, na.rm = TRUE),

    .groups = "drop"
  )

summary_table
```

```{r, echo = FALSE, include=FALSE}
summary_table_report <- summary_table %>%
  mutate(
    across(
      c(
        mean_senior_pct,
        median_senior_pct,
        mean_snap_pct,
        median_snap_pct
      ),
      ~ round(.x, 2)
    )
  )

summary_table_report
```

```{r, echo = FALSE, include=FALSE}
dim(data2015_clean)
names(data2015_clean)
str(data2015_clean)
summary(data2015_clean)
```

```{r, echo = FALSE, include=FALSE}
dim(data2019_clean)
names(data2019_clean)
str(data2019_clean)
summary(data2019_clean)
```

```{r, echo = FALSE, include=FALSE}
colSums(is.na(data2019_clean))
```

```{r, echo = FALSE, include=FALSE}
colSums(is.na(data2015_clean))
```

# Background and Data

**Background and Data**

*Data set and Source of Data*

Our group used the 2015 and 2019 Food Access Research Atlas data sets, available through the U.S. Department of Agriculture Economic Research Service.

*Why does this data set matter?*

The data sets are of interest because they provide insight into access to supermarkets and essential food resources across different communities in the United States of America. They can be used to examine food access among people receiving benefits, those living in areas of high deprivation or low socioeconomic status, those living further away in urban vs rural areas, and different demographic groups. Using data from multiple years also allows us to identify changes over time and determine whether there have been noticeable improvements in food access.

*Types of Data in the data set*

The data sets contain information including census tract, poverty rate, food access at specified distances, and access measures across different racial and age groups, as well as SNAP beneficiaries. Each observation represents a U.S. census tract, with columns containing demographic, socioeconomic and food-access measures associated with that tract. The majority of variables are numerical, including population counts, poverty rates and food-access measures, while categorical variables include state and county. The data sets are cross-sectional snapshots for 2015 and 2019 rather than continuous time-series data. Individual variables represent specific measures of food access for particular demographic groups and distance thresholds.

*Completeness of Data set*

One point of interest was the substantial amount of missing data in the 2019 data set. A number of observations were originally represented as NULL. Although some of these appeared as though they could represent zero values, there was insufficient information to confidently distinguish a true zero from a missing observation. To avoid introducing unsupported assumptions, all NULL and other missing values were therefore converted to NA. This provided a consistent representation of missingness and allowed R to appropriately exclude these observations where necessary during subsequent analysis.

```{r, echo=FALSE}
missing_table <- data.frame(
  Year = c(2015, 2019),
  Senior_Missing = c(
    sum(is.na(data2015_clean$senior_la_pct)),
    sum(is.na(data2019_clean$senior_la_pct))
  ),
  SNAP_Missing = c(
    sum(is.na(data2015_clean$snap_la_pct)),
    sum(is.na(data2019_clean$snap_la_pct))
  )
)

kable(
  missing_table,
  caption = "Number of missing Senior and SNAP low-access share observations, by year.",
  col.names = c("Year", "Senior Missing", "SNAP Missing")
)
```


Missingness was not evenly distributed between the data sets. For the two variables central to our analysis, the 2015 data set contained no missing observations, while the 2019 data set contained 29,953 missing observations for the senior low-access percentage and 29,918 for the SNAP low-access percentage. This substantial difference in completeness between years was therefore considered when interpreting comparisons between 2015 and 2019. Comparisons between the data sets were done with caution. 

```{r, echo=FALSE}
library(dplyr)
library(tibble)
library(knitr)

missing_area_table <- data2019_clean |>
  mutate(
    Area = if_else(urban == 1, "Urban", "Rural")
  ) |>
  group_by(Area) |>
  summarise(
    `Census Tracts` = n(),
    `Senior Missing` = sum(is.na(senior_la_pct)),
    `Senior Missing (%)` = round(mean(is.na(senior_la_pct)) * 100, 1),
    `SNAP Missing` = sum(is.na(snap_la_pct)),
    `SNAP Missing (%)` = round(mean(is.na(snap_la_pct)) * 100, 1),
    .groups = "drop"
  )

kable(
  missing_area_table,
  caption = "Missing Senior and SNAP low-access observations by urban and rural census tracts in 2019.",
  align = c("l", "r", "r", "r", "r", "r")
)
```

Further inspection identified a clear relationship between missingness and urban/rural classification in the 2019 dataset. Although urban census tracts contained a greater number of missing observations overall, this reflects the much larger number of urban tracts in the dataset. Proportionally, 57.4% of rural census tracts had missing values for both the Senior and SNAP low-access measures, compared with 36.2% of urban census tracts. This indicates that missingness is disproportionately concentrated among rural observations and should be considered when interpreting subsequent urban-rural comparisons.

Similarly, both data sets also contained a considerable number of genuine NA values, particularly for SNAP_la_rate (2,405) and senior_la_rate (520).

*Integration of data set*

To enable comparisons between 2015 and 2019, both data sets were cleaned and standardised before integration. Although the data sets had largely consistent structures, variable names were first converted to lowercase to account for minor differences in capitalisation between years. Equivalent variables relevant to our analysis were then selected from each data set and converted to consistent data types. Missing and NULL values were also represented consistently as NA.

Further transformations were required to make the food-access measures directly comparable. An Urban/Rural classification and year indicator were created, while Senior and SNAP low-access measures were constructed using the relevant distance thresholds for urban and rural census tracts. Percentage measures also required standardisation because the 2015 share variables were represented as proportions, while the corresponding 2019 variables were already expressed as percentages. Once the variables were consistently defined and formatted, the data sets could be vertically combined for comparison between years.

```{r, echo=FALSE}
dataset_overview <- tibble(
  Feature = c(
    "Census tracts",
    "Variables retained for analysis",
    "Urban tracts",
    "Rural tracts",
    "Senior access measures",
    "SNAP access measures",
    "Poverty measures"
  ),
  
  `2015` = c(
    nrow(data2015_clean),
    ncol(data2015_clean),
    sum(data2015_clean$urban == 1, na.rm = TRUE),
    sum(data2015_clean$urban == 0, na.rm = TRUE),
    "Yes",
    "Yes",
    "Yes"
  ),
  
  `2019` = c(
    nrow(data2019_clean),
    ncol(data2019_clean),
    sum(data2019_clean$urban == 1, na.rm = TRUE),
    sum(data2019_clean$urban == 0, na.rm = TRUE),
    "Yes",
    "Yes",
    "Yes"
  )
)

kable(
  dataset_overview,
  caption = "Overview of the 2015 and 2019 Food Access Research Atlas datasets.",
  col.names = c("Feature", "2015", "2019"),
  align = c("l", "c", "c")
)
```

\newpage

# Ethics, Privacy & Security

**Ethics**

This project involves data describing vulnerable populations,
specifically seniors aged 65 and over and SNAP-receiving households.
These populations require discretion as careless findings or framing
about them can further stigmatise affected communities. A key
consideration for this analysis is avoiding deficit oriented framing,
practically, this means describing a tract as “low access” reflects a
structural and geographic outcome specifically the distance for a
household to supermarkets or transport infrastructure. It is not
intended to signify the choices of the individuals living in the
household, this is to avoid perpetuating any stereotypes or stigmas. As
a result the language of the report has been mindfully selected to
reflect this. For example we would state “A tract has a high rate of
low-access seniors” instead of language that could be read as describing
a senior or snap receiving households like “seniors in this area don’t
shop at super markets”

Another ethical risk we are considering is an ecological fallacy. All
analysis in this project is conducted at the census-tract level which
uses aggregated rates for an area rather than individual data. So where
a tract is classified as “low access” indicates the aggregate population
meets USDA distance thresholds, it does not mean that every individual
within that tract lacks access. As a result our conclusions are based on
patterns observed in tracts, not on individuals.

**Privacy**

This project uses only publicly available, government-released data
(USDA Food Access Research Atlas and U.S. Census Bureau data), both
provided at the census-tract level. We did no original data collection
of our own, or have any interaction with the individuals involved with
the data. The data is already aggregated tract-areas and released for
public use as verified by the U.S. government so the privacy risk to
individuals is minimal. The risks too to privacy are not technical with
how the data is structured or analysed but in the interpretation of the
data. There is still a risk (although very small risk) of individual
identification for tracts with very small populations e.g a small amount
of senior SNAP households in a rural tract, where in principle it could
be easier to infer information about specific households. This
represents a limitation of the dataset from a privacy perspective,
though it is a low and indirect risk rather than a risk introduced by
our analysis.

**Security**

As mentioned in the privacy section this project uses only public,
pre-aggregated data. The overall security risk associated with this
project is low. We practiced secure data management in this project
where we store all our working data and code within a shared private
Github repository, accessible only to the members of our group as
opposed to a public repository or sharing code in any other external
source. This gave us a controlled and access-limited location for our
data, code and analysis. With some additional time and resources to
allocate into security it would be worthwhile establishing a documented
data deletion plan for after the project concludes.

After harmonisation, the 2015 and 2019 datasets were combined into a
dataset containing 145,395 census-tract observations. Of these, 110,341
were classified as urban and 35,054 as rural, with no missing values in
the urban–rural classification.

\newpage

# Exploratory Data Analysis

**3.1 Overview of the Analysis**

The exploratory analysis focused on the low-supermarket-access shares
for two groups: adults aged 65 years and over, and SNAP-receiving
households. Comparisons were made between 2015 and 2019 and between
urban and rural census tracts. For urban census tracts, low supermarket
access was measured using the one-mile threshold, while the ten-mile
threshold was used for rural census tracts. All low-access shares were
expressed as percentages to ensure consistency between the 2015 and 2019
datasets.Because urban and rural low-accessibility percentages are not
calculated on the same scale, the report's urban-rural comparisons
reflect each area's results under USDA standards—not a direct distance
comparison. This analysis covers 2015–2019 changes and urban-rural tract
differences, with all figures in percentages for consistent
year-over-year comparison.

Table 4 shows a clear urban-rural gap: the mean low-access share for
older adults in urban census tracts is roughly double that in rural
tracts (3.4% vs. 1.7% in 2015; 5.3% vs. 4.0% in 2019), and the same
pattern holds for SNAP households (1.3% vs. 2.9% in 2015; 2.8% vs. 4.3%
in 2019). This pattern reflects the differing USDA thresholds applied to
urban and rural areas, rather than an absolute access gap. The effective
sample size dropped substantially between 2015 and 2019, with rural
tract observations more than halving (from 17,692 to around 7,400),
suggesting that 2019 estimates should be interpreted with greater
caution than those for 2015. Based on the table, every mean exceeds its
corresponding median, with the largest gap observed in the 2015 rural
group (mean 1.7% vs. median 0.0%), indicating a strongly right-skewed
distribution where a small number of tracts with particularly severe
accessibility barriers pull the averages upward.

```{r, echo = FALSE}
summary_table <- combined_data %>%
  group_by(area, year) %>%
  summarise(
    Senior_Mean   = round(mean(senior_la_pct, na.rm = TRUE), 1),
    Senior_Median = round(median(senior_la_pct, na.rm = TRUE), 1),
    Senior_Valid_N = sum(!is.na(senior_la_pct)),
    
    SNAP_Mean     = round(mean(snap_la_pct, na.rm = TRUE), 1),
    SNAP_Median   = round(median(snap_la_pct, na.rm = TRUE), 1),
    SNAP_Valid_N = sum(!is.na(snap_la_pct)),
    n_tracts      = n(),
    .groups = "drop"
  )

kable(summary_table,
      caption = "Mean and median low-access share (%) for seniors and SNAP households, by area and year.",
      col.names = c("Area", "Year", "Senior Mean", "Senior Median", "Senior N (Valid)",
                    "SNAP Mean", "SNAP Median", "SNAP N (Valid)", "N Tracts"))
```

```{r, echo = FALSE, include = FALSE}
missing_table <- data.frame(
  Year = c(2015, 2019),
  Senior_Missing = c(
    sum(is.na(data2015_clean$senior_la_pct)),
    sum(is.na(data2019_clean$senior_la_pct))
  ),
  SNAP_Missing = c(
    sum(is.na(data2015_clean$snap_la_pct)),
    sum(is.na(data2019_clean$snap_la_pct))
  )
)

kable(missing_table,
      caption = "Number of missing Senior and SNAP low-access share observations, by year.",
      col.names = c("Year", "Senior Missing", "SNAP Missing"))
```

**3.2 Senior Low-Access Share: Distribution and Urban–Rural Comparison**

The following sections look at distribution of variables in more detail,
starting with seniors.

Figure 1 shows the distribution of senior low-access shares by year and
urban/rural areas. The distributions were strongly right-skewed across
all four groups, with most census tracts having relatively low senior
low-access shares and a smaller number having much higher values. Urban
census tracts generally showed higher low-access shares than rural
census tracts. The 2019 distributions also appeared shifted toward
higher values than in 2015. However, interpret the 2019 results
cautiously because the senior low-access variable contained substantial
missing data.

```{r, echo = FALSE,fig.width = 5.8, fig.height = 3.0}
ggplot(combined_data, aes(x = senior_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  facet_grid(area ~ year) +
  labs(
    title = "Figure1. Distribution of Senior Low-Access Share",
    x = "Senior Low-Access Share (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Figure 2 provides a more direct comparison between urban and rural
tracts for the senior low-access share. In both 2015 and 2019, the
median senior low-access share was consistently higher in urban tracts
than in rural tracts, rising from 0.6% to 3.5% in urban areas compared
with 0.0% to 1.7% in rural areas. The spread also widened between the
two years: in rural areas, the boxp shifted from being almost flat at
zero in 2015, but by 2019 it had extended noticeably upward by several
percentage points. Urban tracts showed a similar pattern, with the box
stretching further above the median, indicating greater variability
across tracts in 2019. In both years, a number of outlying high values
are visible, which is consistent with the right-skewed distribution
observed in Figure 1.

```{r, echo = FALSE,warning=FALSE, fig.width = 5.8, fig.height = 3.0}
# Senior Boxplot
ggplot(combined_data, 
       aes(x = factor(year), y = senior_la_pct, fill = area)) +
  geom_boxplot(outlier.size = 0.3,na.rm = TRUE) +
  scale_fill_grey() +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = sprintf("%.1f", after_stat(y))),
    position = position_dodge(width = 0.75),
    vjust = +1,
    hjust = +1,
    size = 3,
  ) +
  labs(
    title = "Figure 2. Senior Low-Access Share by Year and Area",
    x = "Year",
    y = "Senior Low-Access Share (%)",
    fill = "Area"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Overall, the senior low-access share was generally higher in urban than
rural census tracts, and the distributions were strongly right-skewed in
both years. The apparent increase from 2015 to 2019 may be partly
attributable to differences in the set of census tracts with available
data, rather than reflecting a genuine change. In 2019, values for low
accessibility among older adults were missing for approximately 30,000
census tracts. This issue is addressed directly in the matched analysis
presented in Section 3.4.

**3.3 SNAP Low-Access Share: Distribution and Urban–Rural Comparison**

This section examines the SNAP low-access share in more detail.

```{r, echo = FALSE,include = FALSE}
ggplot(combined_data, aes(x = snap_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(
    title = "Figure 3: Distribution of SNAP Low-Access Share, All Years and Areas",
    x = "SNAP Low-Access Share (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

Figure 3 displays the distribution of SNAP low-access share by year and
urban-rural regions. The distributions of all four groups exhibit a
strong right-skewed pattern: the SNAP low-access share in most census
areas are relatively low, while only a few census areas have
significantly higher values. Urban census areas show a longer and more
extended tail to the right in both years compared to rural census areas,
indicating greater variability among urban census areas.

```{r, echo = FALSE, fig.width = 5.8, fig.height = 3.0}
ggplot(combined_data,aes(x = snap_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  facet_grid(area ~ year) +
  labs(
    title = "Figure 3. Distribution of SNAP Low-Access Share",
    x = "SNAP Low-Access Share (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Figure 4 shows that the median SNAP low-access share was consistently
higher in urban tracts than in rural tracts in both years, rising from
0.3% to 1.8% in urban areas compared with 0.0% to 1.0% in rural areas.
The spread also widened in both areas: rural boxes expanded from nearly
flush with zero to about one percentage point higher, while urban boxes
stretched further above the median, indicating greater tract-level
variation in 2019. The distribution in both years and for both regions
remained strongly skewed to the right, with high values observed
extending to approximately 85% - 100%, indicating that a small portion
of the surveyed areas were more pronounced in urban areas and continued
to face significantly higher barriers to supermarket accessibility
compared to typical surveyed areas.

```{r, echo = FALSE, warning=FALSE,fig.width = 5.8, fig.height = 3.0}
# SNAP boxplot
ggplot(combined_data, 
       aes(x = factor(year), y = snap_la_pct,fill = area)) +
  geom_boxplot(outlier.size = 0.3, na.rm = TRUE) +
  scale_fill_grey() +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = sprintf("%.1f", after_stat(y))),
    position = position_dodge(width = 0.75),
    vjust = -0.5,
    size = 3
  ) +
  labs(
    title = "Figure 4. SNAP Low-Access Share by Year and Area",
    x = "Year",
    y = "SNAP Low-Access Share (%)",
    fill = "Area"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

In short, the SNAP low-access share was generally higher in urban than
rural census tracts, and the distributions were strongly right-skewed in
both years. The apparent increase from 2015 to 2019 may be partly
attributable to differences in the set of census tracts with available
data, rather than reflecting a genuine change. In 2019, values for SNAP
low-access were missing for approximately 29,918 census tracts. This
issue is addressed directly in the matched analysis presented in Section
3.4.

**3.4 Changes between 2015 and 2019**

To examine changes between 2015 and 2019 more directly, census tracts in
the two datasets were matched using their Census Tract identifiers. A
total of 72,506 census tracts were successfully matched across the two
years. Of these, 21 census tracts changed their urban/rural
classification between 2015 and 2019. These tracts were excluded from
the change analysis because the low-access distance threshold differs
between urban and rural areas. This resulted in 72,485 census tracts
with a stable urban/rural classification.

```{r, echo = FALSE, include = FALSE}
# Match the same census tract in 2015 and 2019, and calculate the changes in the share of seniors and SNAP low-access individuals between the two years.

matched_data <- data2015_clean %>%
  select(
    censustract,
    area_2015 = area,
    senior_pct_2015 = senior_la_pct,
    snap_pct_2015 = snap_la_pct
  ) %>%
  inner_join(
    data2019_clean %>%
      select(
        censustract,
        area_2019 = area,
        senior_pct_2019 = senior_la_pct,
        snap_pct_2019 = snap_la_pct
      ),
    by = "censustract"
  ) %>%
  mutate(
    senior_change = senior_pct_2019 - senior_pct_2015,
    snap_change = snap_pct_2019 - snap_pct_2015
  )

matched_data
```

```{r, echo = FALSE, include = FALSE}
dim(matched_data)
sum(!is.na(matched_data$senior_change))
sum(!is.na(matched_data$snap_change))
```

```{r, echo = FALSE, include = FALSE}
# Senior Change (using all matched tracts, before excluding reclassified tracts)
ggplot(
  matched_data,
  aes(x = area_2015, y = senior_change)) +
  geom_boxplot(na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Change in Senior Low-Access Share, 2015-2019",
    x = "Area",
    y = "Change in percentage points"
  ) +
  theme_minimal()
```

```{r, echo = FALSE, include = FALSE}
# Check the boxplot why is the line 
matched_data %>%
  group_by(area_2015) %>%
  summarise(n = n(), 
            Q1 = quantile(senior_change, 0.25, na.rm = TRUE),
            Median = median(senior_change, na.rm = TRUE),
            Q3 = quantile(senior_change, 0.75, na.rm = TRUE),
            Min = min(senior_change, na.rm = TRUE),
            Max = max(senior_change, na.rm = TRUE),
            pct_exact_zero =mean(senior_change == 0,na.rm = TRUE) * 100
  )

```

```{r, echo = FALSE, include = FALSE}
matched_data %>%
  group_by(area_2015) %>%
  summarise(
    pct_near_zero =
      mean(
        abs(senior_change) < 1e-8,
        na.rm = TRUE
      ) * 100
  )
```

```{r, echo = FALSE, include = FALSE}
# SNAP
ggplot(
  matched_data,
  aes(x = area_2015, y = snap_change)) +
  geom_boxplot(na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Change in SNAP Low-Access Share, 2015-2019",
    x = "Area",
    y = "Change in percentage points"
  ) +
  theme_minimal()
```

```{r, echo = FALSE, include = FALSE}
matched_stable <- matched_data %>%
  filter(area_2015 == area_2019)

matched_stable
```

For the paired analysis, only census tracts with non-missing low-access
shares in both 2015 and 2019 were included. Of the 72,485 tracts with stable 
classification, 42,554 had complete senior data in both years and 42,589 had 
complete SNAP data in both years (Table 5). Change was calculated as:
`Low-access share2019 - Low-access share2015` and is reported in
percentage points. Positive values therefore represent an increase in
low-access share between 2015 and 2019, while negative values represent
a decrease.

```{r, echo = FALSE, include = FALSE}
change_summary <- matched_data %>%
  group_by(area_2015) %>%
  summarise(
    n = sum(!is.na(senior_change)),
    median_change =median(senior_change, na.rm = TRUE),
    mean_change =mean(senior_change, na.rm = TRUE),
    pct_increase =mean(senior_change > 0.1, na.rm = TRUE) * 100,
    pct_decrease =mean(senior_change < -0.1, na.rm = TRUE) * 100,
    pct_stable = mean(abs(senior_change) <= 0.1, na.rm = TRUE) * 100,
    .groups = "drop"
  )

change_summary
```

```{r, echo = FALSE, include = FALSE}
table(
  matched_data$area_2015,
  matched_data$area_2019
)
```

```{r, echo = FALSE, include = FALSE}
# Make sure the number of row between two datasets
nrow(matched_data)
nrow(matched_stable)

nrow(matched_data) - nrow(matched_stable)
```

```{r, echo = FALSE, include = FALSE}
matched_stable <- matched_stable %>%
  mutate(snap_change = snap_pct_2019 - snap_pct_2015)

matched_stable
```

```{r, echo = FALSE, include = FALSE}
senior_paired_data <- matched_stable %>%
  filter(
    !is.na(senior_pct_2015),
    !is.na(senior_pct_2019)
  )

senior_paired <- senior_paired_data %>%
  group_by(area_2015) %>%
  summarise(
    N = n(),
    
    Mean_2015 = mean(senior_pct_2015),
    Mean_2019 = mean(senior_pct_2019),
    
    Median_2015 = median(senior_pct_2015),
    Median_2019 = median(senior_pct_2019),
    
    Mean_Change = mean(senior_change),
    Median_Change = median(senior_change),
    
    .groups = "drop"
  ) %>%
  mutate(
    across(
      c(
        Mean_2015,
        Mean_2019,
        Median_2015,
        Median_2019,
        Mean_Change,
        Median_Change
      ),
      ~ round(.x, 2)
    )
  )

kable(
  senior_paired,
  caption = "Paired summary of Senior low-access share, 2015-2019",
  col.names = c(
    "Area",
    "N",
    "Mean 2015 (%)",
    "Mean 2019 (%)",
    "Median 2015 (%)",
    "Median 2019 (%)",
    "Mean Change (pp)",
    "Median Change (pp)"
  )
)
```

```{r, echo = FALSE, include = FALSE}
snap_paired_data <- matched_stable %>%
  filter(
    !is.na(snap_pct_2015),
    !is.na(snap_pct_2019)
  )

snap_paired <- snap_paired_data %>%
  group_by(area_2015) %>%
  summarise(
    N = n(),
    
    Mean_2015 = mean(snap_pct_2015),
    Mean_2019 = mean(snap_pct_2019),
    
    Median_2015 = median(snap_pct_2015),
    Median_2019 = median(snap_pct_2019),
    
    Mean_Change = mean(snap_change),
    Median_Change = median(snap_change),
    
    .groups = "drop"
  ) %>%
  mutate(
    across(
      c(
        Mean_2015,
        Mean_2019,
        Median_2015,
        Median_2019,
        Mean_Change,
        Median_Change
      ),
      ~ round(.x, 2)
    )
  )

kable(
  snap_paired,
  caption = "Paired summary of SNAP low-access share, 2015-2019",
  col.names = c(
    "Area",
    "N",
    "Mean 2015 (%)",
    "Mean 2019 (%)",
    "Median 2015 (%)",
    "Median 2019 (%)",
    "Mean Change (pp)",
    "Median Change (pp)"
  )
)
```

Table 5 reveals an interesting contrast: the trends for senior and SNAP
households moved in opposite directions between 2015 and 2019. The mean
proportion of older adults rose slightly in both rural (from 3.94% to
3.96%) and urban (from 5.21% to 5.32%) census tracts, whilst the mean
proportion of SNAP recipients fell slightly in both rural (from 2.91% to
2.77%) and urban (from 4.40% to 4.33%) census tracts. These opposite
trends suggest that changes in supermarket accessibility for these two
groups are not driven by the same underlying trend, indicating that the
factors influencing changes in accessibility for older people may not be
entirely the same as those affecting SNAP households. However, the
magnitude of these changes is small relative to the degree of variation
within each group shown in Table 4, and the median changes for all four
groups are exactly zero.

```{r, echo = FALSE}
paired_summary <- bind_rows(
  senior_paired %>%
    mutate(Group = "Senior"),
  
  snap_paired %>%
    mutate(Group = "SNAP")
) %>%
  select(
    Group,
    area_2015,
    N,
    Mean_2015,
    Mean_2019,
    Median_2015,
    Median_2019,
    Mean_Change,
    Median_Change
  ) %>%
  mutate(across(c(Mean_2015, Mean_2019, Median_2015, Median_2019,
                Mean_Change, Median_Change),
              ~ {
                x <- round(.x, 2)
                x[x == 0] <- 0
                sprintf("%.2f", x)
              })) %>%
  arrange(Group, area_2015)

kable(
  paired_summary,
  caption = "Paired summary of Senior and SNAP low-access shares, 2015-2019",
  col.names = c(
    "Group",
    "Area",
    "N",
    "Mean 2015 (%)",
    "Mean 2019 (%)",
    "Median 2015 (%)",
    "Median 2019 (%)",
    "Mean Change (pp)",
    "Median Change (pp)"
  ),  
  align = c(
    "c", "l",
    "r", "r", "r", "r", "r", "r", "r"
  )
)
```

```{r, echo=FALSE, include=FALSE}
#matched_stable %>%
#  group_by(area_2015) %>%
#  summarise(
#    n = sum(!is.na(snap_change)),
#    Q1 = quantile(snap_change, 0.25, na.rm = TRUE),
#    Median = median(snap_change, na.rm = TRUE),
#    Q3 = quantile(snap_change, 0.75, na.rm = TRUE),
#    Min = min(snap_change, na.rm = TRUE),
#    Max = max(snap_change, na.rm = TRUE),
#    .groups = "drop"
#  )

#matched_stable %>%
#  group_by(area_2019) %>%
#  summarise(
#    n = sum(!is.na(snap_change)),
#    Q1 = quantile(snap_change, 0.25, na.rm = TRUE),
#    Median = median(snap_change, na.rm = TRUE),
#    Q3 = quantile(snap_change, 0.75, na.rm = TRUE),
#    Min = min(snap_change, na.rm = TRUE),
#    Max = max(snap_change, na.rm = TRUE),
#    .groups = "drop"
#  )

snap_change_summary <- matched_stable %>%
  group_by(Area = area_2015) %>%
  summarise(
    N = sum(!is.na(snap_change)),
    Q1 = round(quantile(snap_change, 0.25, na.rm = TRUE), 2),
    Median = round(median(snap_change, na.rm = TRUE), 2),
    Q3 = round(quantile(snap_change, 0.75, na.rm = TRUE), 2),
    Min = round(min(snap_change, na.rm = TRUE), 2),
    Max = round(max(snap_change, na.rm = TRUE), 2),
    .groups = "drop"
  )

kable(
  snap_change_summary,
  caption = "Summary of changes in SNAP low-access share, 2015–2019, by area"
)
```

```{r,echo=FALSE,include=FALSE, fig.width = 5.8, fig.height = 3.0}
ggplot(matched_data,aes(x = senior_change)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  facet_wrap(~ area_2015) +
  geom_vline(xintercept = 0,linetype = "dashed") +
  labs(
    title = "Distribution of Changes in Senior Low-Access Share, 2015-2019 (unfiltered)",
    x = "Change in percentage points",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Figure 5 shows that changes in senior low-access share were strongly
concentrated around zero in both rural and urban census tracts. This is
consistent with the zero median changes reported in Table 5. Although
most census tracts experienced relatively small changes, a smaller
number showed substantially larger positive or negative changes.

```{r,echo=FALSE, fig.width = 5.8, fig.height = 3.0, out.width="80%"}
ggplot(matched_stable, aes(x = senior_change)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  facet_wrap(~ area_2015) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Figure 5. Distribution of Changes in Senior Low-Access Share (2015-2019)",
    x = "Change in percentage points",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Figure 6 shows that change distributions in SNAP Low-Access share were
also strongly concentrated around zero, although some census tracts
experienced larger increases or decreases. This is consistent with the
zero median changes reported in Table 5, and together they indicate that
the overall changes between the two years were small for most matched
census tracts.

```{r,echo=FALSE, fig.width = 5.8, fig.height = 3.0, out.width="80%"}
ggplot(matched_stable,aes(x = snap_change)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  facet_wrap(~ area_2015) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Figure 6. Distribution of Changes in SNAP Low-Access Share (2015-2019)",
    x = "Change in percentage points",
    y = "Number of Census Tracts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

```{r,echo=FALSE,include=FALSE}
matched_stable %>%
  group_by(area_2015) %>%
  summarise(
    n = sum(!is.na(snap_change)),
    Q1 = quantile(snap_change,0.25,na.rm = TRUE),
    Median = median(snap_change,na.rm = TRUE),
    Q3 = quantile(snap_change,0.75,na.rm = TRUE),
    Min = min(snap_change,na.rm = TRUE),
    Max = max(snap_change,na.rm = TRUE),
    pct_near_zero =mean(abs(snap_change) < 1e-8,na.rm = TRUE) * 100,
    .groups = "drop"
  )
```

In summary, the paired analysis illustrates that the changes at the
census tract level between 2015 and 2019 were generally small. Senior
and SNAP low-access shares moved in opposite directions on average, but
median changes were zero across all four groups, indicating that the
small mean shifts were driven by a minority of tracts with larger
changes rather than a broad trend. These findings are descriptive
results and do not establish statistical significance; the paired
analysis itself cannot eliminate the potential bias caused by missing
data. If there are systematic differences between the census tracts with
missing 2019 data and those with complete data, the above paired results
may still be affected.

**3.5 Additional Comparison: Mean Low-Access Share by Group**

To talk about... median much lower then means = skewed data. Urban
vastly outnumber rural. Values increase across all four rows

keeping in mind a rural tract requires households and seniors to be at
least 10 miles away compared to a min of 1

```{r, echo=FALSE, fig.width=5.8, fig.height=3.0, out.width="90%"}
library(tidyr)

dot_data <- summary_table %>%
  select(area, year, Senior_Mean, SNAP_Mean) %>%
  pivot_longer(cols = c(Senior_Mean, SNAP_Mean),
               names_to = "Group", values_to = "Mean_Rate") %>%
  mutate(Group = recode(Group, Senior_Mean = "Seniors", SNAP_Mean = "SNAP"))

ggplot(dot_data, aes(x = Mean_Rate, y = Group, color = factor(year))) +
  geom_point(size = 3) +
  facet_wrap(~area) +
  labs(
    title = "Figure 7. Mean Low-Access Share by Group, Area, and Year",
    x = "Mean Low-Access Share (%)",
    y = NULL,
    color = "Year"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

Across both Senior and SNAP low-access shares, 2019 shows a higher mean
rate than 2015, and Urban tracts show a higher mean rate than Rural
tracts, in both years and for both groups

Because "low-access" is defined using a much stricter distance threshold
in urban areas (1 mile) than rural areas (10 miles), a higher urban
low-access rate does not necessarily mean urban residents travel further
to reach a supermarket than rural residents. This may just reflect that
urban areas are held to a stricter access standard. A rural resident
traveling several miles to a store may still count as "having access"
under the rural threshold, while a much shorter distance in a city would
count as "low access."

As a result, the lower mean low-access share in rural areas should not
be read as evidence that rural food access is genuinely better. Rural
residents may need to travel considerably farther in absolute terms
before being classified as low-access at all, meaning the rural rate
likely understates the true burden of distance compared to the urban
rate. This distinction should be kept in mind whenever urban and rural
low-access rates are compared directly in this report, rather than
treating them as measuring the same underlying concept.

**3.6 Summary**

In conclusion, the exploratory analysis has identified several important
patterns in the indicators of senior and SNAP low-access. First, both
low-access share distributions were strongly right-skewed. Most census
tracts had relatively low low-access shares, while a smaller number of
tracts had substantially higher values. Thus, the mean was generally
higher than the median, indicating that the mean was influenced by
census tracts with relatively large low-access shares.

Across both measures, urban census tracts consistently showed higher
low-access shares than rural tracts in both years. However, since the
urban census tracts use a stricter one-mile threshold while the rural
census tracts use a ten-mile threshold, this disparity reflects the
classification results of each region based on their own USDA standards,
rather than directly measuring which group of residents is actually
farther from the supermarket.

Although cross-sectional distributions in 2019 appeared to be higher
than those in 2015, such comparisons should be interpreted with caution
as there were a large number of missing values in the 2019 data.The
paired analysis provided a more direct comparison of the same census
tracts across years and showed that changes were generally small. Senior
low-access shares increased slightly on average, while SNAP low-access
shares decreased slightly, and the median change was zero across both
rural and urban census tracts.

\newpage

# Individual Contributions
