---
title: "FoodAccessPREP"
author: "Oliver Temel"
date: "2026-08-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Background and Data**

Our group used the 2015 and 2019 Food Access Research Atlas datasets, available through the U.S. Department of Agriculture Economic Research Service.

The datasets are of interest because they provide insight into access to supermarkets and essential food resources across different communities. They can be used to examine food access among people receiving benefits, those living in areas of high deprivation or low socioeconomic status, those living further away in urban vs rural areas, and different demographic groups. Using data from multiple years also allows us to identify changes over time and determine whether there have been noticeable improvements in food access.

The datasets contain information including census tract, poverty rate, food access at specified distances, and access measures across different racial and age groups, as well as SNAP beneficiaries. The majority of the variables are numerical, with some categorical variables such as census tract, state, and county. The data is structured so that individual variables represent specific measures of food access for particular demographic groups and distance thresholds.

One point of interest was that the 2019 dataset contained a large number of NULL values. These were assumed to represent zero values, as many of the affected variables represented whether an observation belonged to a particular demographic or access category. For example, a variable could represent the number of Native American individuals experiencing low food access at 20 miles, where a NULL indicated that there were no individuals meeting that condition. In contrast, both datasets also contained a considerable number of genuine NA values, particularly for SNAP_la_rate (2,405) and senior_la_rate (520).

To enable comparisons between years, our group used both the 2015 and 2019 Food Access Research Atlas datasets. The datasets were straightforward to integrate because they contained consistent and largely homogeneous variable names. We therefore retained the same variables from each dataset, allowing us to make meaningful comparisons in food access between 2015 and 2019.

## Ethics, Privacy & Security

**Ethics**

This project involves data describing vulnerable populations, specifically seniors aged 65 and over and SNAP-receiving households. These populations require discretion as careless findings or framing about them can further stigmatise affected communities. A key consideration for this analysis is avoiding deficit oriented framing, practically, this means describing a tract as “low access” reflects a structural and geographic outcome specifically the distance for a household to supermarkets or transport infrastructure. It is not intended to signify the choices of the individuals living in the household, this is to avoid perpetuating any stereotypes or stigmas. As a result the language of the report has been mindfully selected to reflect this. For example we would state “A tract has a high rate of low-access seniors” instead of language that could be read as describing  a senior or snap receiving households like “seniors in this area don’t shop at super markets” 

Another ethical risk we are considering is an ecological fallacy. All analysis in this project is conducted at the census-tract level which uses aggregated rates for an area rather than individual data. So where a tract is classified as “low access” indicates the aggregate population meets USDA distance thresholds, it does not mean that every individual within that tract lacks access. As a result our conclusions are based on patterns observed in tracts, not on individuals. 

**Privacy**

This project uses only publicly available, government-released data (USDA Food Access Research Atlas and U.S. Census Bureau data), both provided at the census-tract level. We did no original data collection of our own, or have any interaction with the individuals involved with the data. The data is already aggregated tract-areas and released for public use as verified by the U.S. government so the privacy risk to individuals is minimal. The risks too to privacy are not technical with how the data is structured or analysed but in the interpretation of the data. There is still a risk (although very small risk) of individual identification for tracts with very small populations e.g a small amount of senior SNAP households in a rural tract, where in principle it could be easier to infer information about specific households. This represents a limitation of the dataset from a privacy perspective, though it is a low and indirect risk rather than a risk introduced by our analysis.

**Security**

As mentioned in the privacy section this project uses only public, pre-aggregated data. The overall security risk associated with this project is low. We practiced secure data management in this project where we store all our working data and code within a shared private Github repository, accessible only to the members of our group as opposed to a public repository or sharing code in any other external source. This gave us a controlled and access-limited location for our data, code and analysis. With some additional time and resources to allocate into security it would be worthwhile establishing a documented data deletion plan for after the project concludes. 

# DATA 2015

```{r Data Inputs 2015}
library(readxl)
DATA2015 <- read_excel("FoodAccessResearchAtlasData2015.xlsx", 
                       sheet = 3, na = c("", "NULL"))
head(DATA2015)
```

# DATA 2019

```{r DATA Inputs 2019}
DATA2019 <- read_excel("FoodAccessResearchAtlasData2019.xlsx", 
                       sheet = 3, na = c("", "NULL"))
head(DATA2019)
```

```{r}
dim(DATA2015)
dim(DATA2019)

names(DATA2015)
names(DATA2019)

str(DATA2015)
str(DATA2019)

```

```{r}
# To Unify variable names
names(DATA2015) <- tolower(names(DATA2015))
names(DATA2019) <- tolower(names(DATA2019))
```


```{r}
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
```

```{r}
# Select variables we need
data2015_clean <- DATA2015 %>%
  select(
    censustract,
    state,
    county,
    urban,
    pop2010,
    ohu2010,
    laseniors1,
    laseniors10,
    lasnap1,
    laseniors1share,
    laseniors10share,
    lasnap1share,
    lasnap10share,
    lasnap10,
    laseniors20,
    laseniorshalf,
    laseniorshalfshare,
    tractseniors,
    tractsnap,
  )

data2019_clean <- DATA2019 %>%
  select(
    censustract,
    state,
    county,
    urban,
    pop2010,
    ohu2010,
    laseniors1,
    laseniors10,
    laseniors1share,
    laseniors10share,
    lasnap1share,
    lasnap10share,
    lasnap1,
    lasnap10,
    laseniors20,
    laseniorshalf,
    laseniorshalfshare,
    tractseniors,
    tractsnap, 
)
```

```{r}
numeric_vars <- c(
  "urban",
  "pop2010",
  "ohu2010",
  "laseniors1",
  "laseniors10",
  "laseniors1share",
  "laseniors10share",
  "lasnap1",
  "lasnap10",
  "lasnap1share",
  "lasnap10share",
  "tractseniors",
  "tractsnap"
)

data2015_clean <- data2015_clean %>%
  mutate(
    across(
      any_of(numeric_vars),
      as.numeric
    )
  )

data2019_clean <- data2019_clean %>%
  mutate(
    across(
      any_of(numeric_vars),
      as.numeric
    )
  )

combined_data <- bind_rows(
  data2015_clean,
  data2019_clean
)

combined_data
```

```{r}
sapply(
  data2015_clean[numeric_vars],
  class
)

sapply(
  data2019_clean[numeric_vars],
  class
)
```

```{r add year}
data2015_clean <- data2015_clean %>%
  mutate(year = 2015)

data2019_clean <- data2019_clean %>%
  mutate(year = 2019)
```

```{r}
#Create Urban/Rural
data2015_clean <- data2015_clean %>%
  mutate(
    year = 2015,
    area = if_else(
      urban == 1,
      "Urban",
      "Rural",
      missing = NA_character_
    )
  )

data2019_clean <- data2019_clean %>%
  mutate(
    year = 2019,
    area = if_else(
      urban == 1,
      "Urban",
      "Rural",
      missing = NA_character_
    )
  )
```

```{r}
# 2015 of senior and snap
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_n = case_when(
      urban == 1 ~ laseniors1,
      urban == 0 ~ laseniors10,
      TRUE ~ NA_real_
    ),

    snap_la_n = case_when(
      urban == 1 ~ lasnap1,
      urban == 0 ~ lasnap10,
      TRUE ~ NA_real_
    )
  )

# 2019 of senior and snap
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_n = case_when(
      urban == 1 ~ laseniors1,
      urban == 0 ~ laseniors10,
      TRUE ~ NA_real_
    ),

    snap_la_n = case_when(
      urban == 1 ~ lasnap1,
      urban == 0 ~ lasnap10,
      TRUE ~ NA_real_
    )
  )
```

```{r}
# Show precentage
# 2015
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_pct = case_when(
      urban == 1 ~ laseniors1share * 100,
      urban == 0 ~ laseniors10share * 100,
      TRUE ~ NA_real_
    ),

    snap_la_pct = case_when(
      urban == 1 ~ lasnap1share * 100,
      urban == 0 ~ lasnap10share * 100,
      TRUE ~ NA_real_
    )
  )

# 2019
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_pct = case_when(
      urban == 1 ~ laseniors1share,
      urban == 0 ~ laseniors10share,
      TRUE ~ NA_real_
    ),

    snap_la_pct = case_when(
      urban == 1 ~ lasnap1share,
      urban == 0 ~ lasnap10share,
      TRUE ~ NA_real_
    )
  )
```


```{r}
# Seniors
data2015_clean <- data2015_clean %>%
  mutate(
    senior_la_n = if_else(
      urban == 1,
      laseniors1,
      laseniors10
    )
  )

# SNAP
data2015_clean <- data2015_clean %>%
  mutate(
    snap_la_n = if_else(
      urban == 1,
      lasnap1,
      lasnap10
    )
  )

# Seniors
data2019_clean <- data2019_clean %>%
  mutate(
    senior_la_n = if_else(
      urban == 1,
      laseniors1,
      laseniors10
    )
  )

# SNAP
data2019_clean <- data2019_clean %>%
  mutate(
    snap_la_n = if_else(
      urban == 1,
      lasnap1,
      lasnap10
    )
  )
```

```{r}
summary(data2015_clean)
summary(data2019_clean)
```

```{r}
#data2019_clean <- data2019_clean %>%
#  mutate(
#    across(
#      c(laseniors1, laseniors10, lasnap1, lasnap10,
#        laseniors20,laseniorshalf,laseniorshalfshare,senior_la_n, snap_la_n),
#      ~ as.numeric(ifelse(.x == "NULL", "0", .x))
#    )
#  )
```

```{r}
summary(data2019_clean)
```

```{r}
library(knitr)

dataset_table <- data.frame(
  Dataset = c("2015 Food Access Research Atlas",
              "2019 Food Access Research Atlas"),
  Observations = c(72864, 72531),
  Variables = c(147, 147),
  Unit_of_Analysis = c("Census tract", "Census tract"),
  Main_Data_Types = c(
    "Geographic identifiers, urban/rural indicator, population counts, food-access measures",
    "Geographic identifiers, urban/rural indicator, population counts, food-access measures"
  )
)

kable(
  dataset_table,
  caption = "Structure of the 2015 and 2019 Food Access Research Atlas datasets",
  col.names = c(
    "Dataset",
    "Observations",
    "Variables",
    "Unit of Analysis",
    "Main Data Types"
  ),
  align = c("l", "r", "r", "l", "l")
)
```

```{r}
summary_stats <- bind_rows(
  
  data2015_clean %>%
    transmute(
      year = 2015,
      Senior = senior_la_pct,
      SNAP = snap_la_pct
    ),
  
  data2019_clean %>%
    transmute(
      year = 2019,
      Senior = senior_la_pct,
      SNAP = snap_la_pct
    )
  
) %>%
  
  pivot_longer(
    cols = c(Senior, SNAP),
    names_to = "Group",
    values_to = "Low_Access_Pct"
  ) %>%
  
  group_by(year, Group) %>%
  
  summarise(
    Min = min(Low_Access_Pct, na.rm = TRUE),
    Q1 = quantile(Low_Access_Pct, 0.25, na.rm = TRUE),
    Median = median(Low_Access_Pct, na.rm = TRUE),
    Mean = mean(Low_Access_Pct, na.rm = TRUE),
    Q3 = quantile(Low_Access_Pct, 0.75, na.rm = TRUE),
    Max = max(Low_Access_Pct, na.rm = TRUE),
    Missing = sum(is.na(Low_Access_Pct)),
    .groups = "drop"
  ) %>%
  
  mutate(
    across(
      c(Min, Q1, Median, Mean, Q3, Max),
      ~ round(.x, 2)
    )
  )

summary_stats
```

```{r}
kable(
  summary_stats,
  caption = "Summary Statistics for Senior and SNAP Low-Access Shares, 2015 and 2019",
  col.names = c(
    "Year",
    "Group",
    "Min (%)",
    "Q1 (%)",
    "Median (%)",
    "Mean (%)",
    "Q3 (%)",
    "Max (%)",
    "Missing"
  ),
  align = c(
    "c", "l",
    "r", "r", "r", "r", "r", "r", "r"
  )
)
```

```{r}
# Check whether over 100
data2015_clean %>%
  summarise(
    senior_over_100 = sum(senior_la_pct > 100, na.rm = TRUE),
    snap_over_100 = sum(snap_la_pct > 100, na.rm = TRUE)
  )

data2019_clean %>%
  summarise(
    senior_over_100 = sum(senior_la_pct > 100, na.rm = TRUE),
    snap_over_100 = sum(snap_la_pct > 100, na.rm = TRUE)
  )
```
```{r}
data2015_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )

data2019_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )
```

```{r}
options(digits = 15)

data2015_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )

data2019_clean %>%
  filter(snap_la_pct > 100) %>%
  select(
    censustract,
    urban,
    area,
    lasnap1share,
    lasnap10share,
    snap_la_pct
  )
```

```{r}
# Show precentage
#data2015_clean <- data2015_clean %>%
#  mutate(
#    senior_la_pct = senior_la_rate * 100,
#    snap_la_pct = snap_la_rate * 100
#  )

#data2019_clean <- data2019_clean %>%
#  mutate(
#    senior_la_pct = senior_la_rate * 100,
#    snap_la_pct = snap_la_rate * 100
#  )
```


```{r}
combined_data <- bind_rows(
  data2015_clean,
  data2019_clean
)
combined_data
```
```{r}
dim(combined_data)
table(combined_data$year)
table(combined_data$area, useNA = "ifany")
```
Explain:
After harmonisation, the 2015 and 2019 datasets were combined into a dataset containing 145,395 census-tract observations. Of these, 110,341 were classified as urban and 35,054 as rural, with no missing values in the urban–rural classification.

```{r}
summary_table <- combined_data %>%
  group_by(year, area) %>%
  summarise(
    n_tracts = n(),

    n_valid_senior =
      sum(!is.na(senior_la_pct)),

    n_valid_snap =
      sum(!is.na(snap_la_pct)),

    mean_senior_pct =
      mean(senior_la_pct, na.rm = TRUE),

    median_senior_pct =
      median(senior_la_pct, na.rm = TRUE),

    mean_snap_pct =
      mean(snap_la_pct, na.rm = TRUE),

    median_snap_pct =
      median(snap_la_pct, na.rm = TRUE),

    .groups = "drop"
  )
#summary_table <- combined_data %>%
#  group_by(year, area) %>%
#  summarise(n_tracts = n(),
#    senior_la_rate =
#      sum(senior_la_n, na.rm = TRUE) /
#      sum(tractseniors, na.rm = TRUE),
#    snap_la_rate =
#      sum(snap_la_n, na.rm = TRUE) /
#      sum(tractsnap, na.rm = TRUE),
#    .groups = "drop"
#  ) %>%

#  mutate(
#    senior_la_pct =
#      senior_la_rate * 100,
#    snap_la_pct =
#      snap_la_rate * 100
# )

summary_table
```

```{r}
summary_table_report <- summary_table %>%
  mutate(
    across(
      c(
        mean_senior_pct,
        median_senior_pct,
        mean_snap_pct,
        median_snap_pct
      ),
      ~ round(.x, 2)
    )
  )

summary_table_report
```

```{r}
dim(data2015_clean)
```

```{r}
names(data2015_clean)
```

```{r}
str(data2015_clean)
```

```{r}
summary(data2015_clean)
```

Data Cleaning and Missing Data.

```{r}
dim(data2019_clean)
```

```{r}
names(data2019_clean)
```

```{r}
str(data2019_clean)
```

```{r}
summary(data2019_clean)
```

```{r}
colSums(is.na(data2019_clean))
```

```{r}
dim(data2015_clean)
```

```{r}
names(data2015_clean)
```

```{r}
str(data2015_clean)
```

```{r}
summary(data2015_clean)
```

```{r}
colSums(is.na(data2015_clean))
```

```{r}
#summary(data2015_clean$snap_la_rate)
#summary(data2019_clean$snap_la_rate)
```

```{r}
#summary(data2015_clean$senior_la_rate)
#summary(data2019_clean$senior_la_rate)
```


```{r}
#data2015_clean %>%
#  summarise(
#    n_over_1 = sum(snap_la_rate > 1, na.rm = TRUE),
#    max_rate = max(snap_la_rate, na.rm = TRUE)
#  )

#data2019_clean %>%
#  summarise(
#    n_over_1 = sum(snap_la_rate > 1, na.rm = TRUE),
#    max_rate = max(snap_la_rate, na.rm = TRUE)
#  )

```

```{r}
#data2015_clean %>%
#  filter(snap_la_rate > 1) %>%
#  select(
#    censustract,
#    urban,
#    tractsnap,
#    lasnap1,
#    lasnap10,
#    snap_la_n,
#    snap_la_rate
#  ) %>%
#  head(20)

#data2019_clean %>%
#  filter(snap_la_rate > 1) %>%
#  select(
#    censustract,
#    urban,
#    tractsnap,
#    lasnap1,
#    lasnap10,
#    snap_la_n,
#    snap_la_rate
#  ) %>%
#  head(20)

```

```{r}
#data2015_clean %>%
#  summarise(
#    n_numerator_gt_denominator =
#      sum(snap_la_n > tractsnap, na.rm = TRUE)
#  )

#data2019_clean %>%
#  summarise(
#    n_numerator_gt_denominator =
#      sum(snap_la_n > tractsnap, na.rm = TRUE)
#  )

```

```{r}
# Senior Boxplot
ggplot(
  combined_data,
  aes(
    x = factor(year),
    y = senior_la_pct,
    fill = area
  )
) +
  geom_boxplot(
    outlier.size = 0.3,
    na.rm = TRUE
  ) +
  scale_fill_grey() +
  labs(
    title = "Senior Low-Access Share by Year and Area",
    x = "Year",
    y = "Senior Low-Access Share (%)",
    fill = "Area"
  ) +
  theme_minimal()
```

```{r}
# SNAP boxplot
ggplot(
  combined_data,
  aes(
    x = factor(year),
    y = snap_la_pct,
    fill = area
  )
) +
  geom_boxplot(
    outlier.size = 0.3,
    na.rm = TRUE
  ) +
  scale_fill_grey() +
  labs(
    title = "SNAP Low-Access Share by Year and Area",
    x = "Year",
    y = "SNAP Low-Access Share (%)",
    fill = "Area"
  ) +
  theme_minimal()
```

```{r}
matched_data <- data2015_clean %>%
  select(
    censustract,
    area_2015 = area,
    senior_pct_2015 = senior_la_pct,
    snap_pct_2015 = snap_la_pct
  ) %>%
  inner_join(
    data2019_clean %>%
      select(
        censustract,
        area_2019 = area,
        senior_pct_2019 = senior_la_pct,
        snap_pct_2019 = snap_la_pct
      ),
    by = "censustract"
  ) %>%
  mutate(
    senior_change = senior_pct_2019 - senior_pct_2015,
    snap_change = snap_pct_2019 - snap_pct_2015
  )

matched_data
```

# Check how many census tracts matched?
```{r}
dim(matched_data)
```

```{r}
sum(!is.na(matched_data$senior_change))
```

```{r}
sum(!is.na(matched_data$snap_change))
```

```{r}
# Senior Change
ggplot(
  matched_data,
  aes(x = area_2015, y = senior_change)) +
  geom_boxplot(na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Change in Senior Low-Access Share, 2015–2019",
    x = "Area",
    y = "Change in percentage points"
  ) +
  theme_minimal()
```

```{r}
# Check the boxplot why is the line 
matched_data %>%
  group_by(area_2015) %>%
  summarise(n = n(), 
            Q1 = quantile(senior_change, 0.25, na.rm = TRUE),
            Median = median(senior_change, na.rm = TRUE),
            Q3 = quantile(senior_change, 0.75, na.rm = TRUE),
            Min = min(senior_change, na.rm = TRUE),
            Max = max(senior_change, na.rm = TRUE),
            pct_exact_zero =mean(senior_change == 0,na.rm = TRUE) * 100
  )

```
```{r}
matched_data %>%
  group_by(area_2015) %>%
  summarise(
    pct_near_zero =
      mean(
        abs(senior_change) < 1e-8,
        na.rm = TRUE
      ) * 100
  )
```

Reason for why choose histogram to show the Change in Senior Low-Access Share between 2015 and 2019:
Based on the summary, we find the boxplot is right. However, the vast majority of changes in census tracts are very close to zero. 
From the summary, especially Urban, 
Q1 = -4.019 × 10^-14
Median = 0
Q3 = 4.974 × 10^-14
The real meaning of these three number is 0% points. Thus, the boxplot of Urban: Q1 ≈ Median ≈ Q3 ≈ 0, it shows the box was naturally compressed into a line near y=0.

While, Rural is not completely 0: 
Q1 ≈ -0.014
Median = 0
Q3 ≈ 0.052
However, the IQR just around 0.052−(−0.014) ≈ 0.066. There is the very small percentage points.
Compare the Min(-38.16) and Max(+40.72), so the vertical axis is stretched very wide to accommodate these extreme changes, making the box even less visible.

The key finding We cannot be fully trust "pct_exact_zero".
You now obtain:
Rural: 10.41%
Urban: 15.50%
It seems that only 10–15% have remained unchanged.

However, for Urban, Q1 and Q3 are:
-0.000000000000040
+0.000000000000050

This is actually floating-point numerical noise.
R will consider:
0.00000000000005 != 0
So you use:
senior_change == 0

It will classify many observations that are actually equivalent to 0 as "non-zero changes".

```{r}
# SNAP
ggplot(
  matched_data,
  aes(x = area_2015, y = snap_change)) +
  geom_boxplot(na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Change in SNAP Low-Access Share, 2015–2019",
    x = "Area",
    y = "Change in percentage points"
  ) +
  theme_minimal()
```
```{r}
matched_stable %>%
  group_by(area_2015) %>%
  summarise(
    n = sum(!is.na(snap_change)),
    Q1 = quantile(snap_change, 0.25, na.rm = TRUE),
    Median = median(snap_change, na.rm = TRUE),
    Q3 = quantile(snap_change, 0.75, na.rm = TRUE),
    Min = min(snap_change, na.rm = TRUE),
    Max = max(snap_change, na.rm = TRUE),
    .groups = "drop"
  )
```
Still check, Q1, Median, and Q3 all close to 0, it shows the data itself is highly concentrated around 0.

```{r}
library(ggplot2)
ggplot(data2019_clean, aes(x = snap_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(
    title = "Distribution of SNAP Low-Access Share, 2019",
    x = "SNAP Low-Access Share (%)",
    y = "Number of Census Tracts"
  )
```

```{r}
library(ggplot2)
ggplot(data2015_clean, aes(x = snap_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(
    title = "Distribution of SNAP Low-Access Share, 2015",
    x = "SNAP Low-Access Share (%)",
    y = "Number of Census Tracts"
  )
```

```{r}
library(ggplot2)
ggplot(data2019_clean, aes(x = senior_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(
    title = "Distribution of Senior Low-Access Share, 2019",
    x = "Senior Low-Access Share (%)",
    y = "Number of Census Tracts"
  )
```

```{r}
ggplot(data2015_clean, aes(x = senior_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(
    title = "Distribution of Senior Low-Access Share, 2015",
    x = "Senior Low-Access Share (%)",
    y = "Number of Census Tracts"
  )
```


```{r}
ggplot(combined_data, aes(x = senior_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  facet_grid(area ~ year) +
  labs(
    title = "Distribution of Senior Low-Access Share",
    x = "Senior Low-Access Share (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

```{r}
ggplot(combined_data,aes(x = snap_la_pct)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  facet_grid(area ~ year) +
  labs(
    title = "Distribution of SNAP Low-Access Share",
    x = "SNAP Low-Access Share (%)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

```{r}
ggplot(
  matched_data,
  aes(x = senior_change)
) +
  geom_histogram(
    bins = 50,
    na.rm = TRUE
  ) +
  facet_wrap(
    ~ area_2015
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Distribution of Changes in Senior Low-Access Share, 2015–2019",
    x = "Change in percentage points",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```
This will show more clearly than boxplot:
A large number of tracts are concentrated around 0;
Which ones are increases?
Which ones are reductions?
Is the distribution of Urban and Rural areas different?

```{r}
change_summary <- matched_data %>%
  group_by(area_2015) %>%
  summarise(
    n = sum(!is.na(senior_change)),
    
    median_change =
      median(senior_change, na.rm = TRUE),
    
    mean_change =
      mean(senior_change, na.rm = TRUE),
    
    pct_increase =
      mean(senior_change > 0.1, na.rm = TRUE) * 100,
    
    pct_decrease =
      mean(senior_change < -0.1, na.rm = TRUE) * 100,
    
    pct_stable =
      mean(abs(senior_change) <= 0.1, na.rm = TRUE) * 100,
    
    .groups = "drop"
  )

change_summary
```
```{r}
table(
  matched_data$area_2015,
  matched_data$area_2019
)
```
```{r}
matched_stable1 <- matched_data %>%
  filter(area_2015 == area_2019)

matched_stable1
```

```{r}
ggplot(
  matched_stable,
  aes(x = senior_change)
) +
  geom_histogram(
    bins = 50,
    na.rm = TRUE
  ) +
  facet_wrap(~ area_2015) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Distribution of Changes in Senior Low-Access Share, 2015–2019",
    x = "Change in percentage points",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```
We should use the second which better than last one. Even though we cannot see observable difference, it shows the results which checked: Rural → Urban：10; Urban → Rural：11 .
There are 21 census tracts deleted.

Therefore, although the second graph does not show any visual difference, it is more rigorous in its approach because it ensures that:
For rural areas, the comparison always uses a 10-mile measure, and for urban areas, the comparison always uses a 1-mile measure. 
So, the subsequent 2015–2019 change analysis suggests using "matched_stable".


```{r}
# Make sure the number of row between two datasets
nrow(matched_data)
nrow(matched_stable)

nrow(matched_data) - nrow(matched_stable)
```


```{r}
matched_stable <- matched_stable %>%
  mutate(snap_change = snap_pct_2019 - snap_pct_2015)

matched_stable
```

```{r}
ggplot(matched_stable,aes(x = snap_change)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  facet_wrap(~ area_2015) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Distribution of Changes in SNAP Low-Access Share, 2015–2019",
    x = "Change in percentage points",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

```{r}
matched_stable %>%
  group_by(area_2015) %>%
  summarise(
    n = sum(!is.na(snap_change)),
    Q1 = quantile(snap_change,0.25,na.rm = TRUE),
    Median = median(snap_change,na.rm = TRUE),
    Q3 = quantile(snap_change,0.75,na.rm = TRUE),
    Min = min(snap_change,na.rm = TRUE),
    Max = max(snap_change,na.rm = TRUE),
    pct_near_zero =mean(abs(snap_change) < 1e-8,na.rm = TRUE) * 100,
    .groups = "drop"
  )
```
Then, we can compare:
Senior change: Rural vs Urban
SNAP change: Rural vs Urban

